---
title: "Chapter 1 robustness checks and supplementary analyses"
author: "Helen E. Payne"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

This file implements the methodological corrections and robustness
checks identified in the Chapter 1 review:

1. Boundary-corrected G×Y likelihood-ratio tests (p-values halved)
2. Transect-weighted population mean fitness W̄ (instead of W̄ at
   Transect = 0)
3. Parametric bootstrap SE for VA(W), demonstrated on the BB Year-2 fit
   (the case where the delta-method SE is most likely to be
   non-conservative)
4. HR 2022 perfect-separation sensitivity: equivalence of 5-node and
   3-node W̄ estimates
5. Quantitative predicted-vs-realized rate-of-adaptation comparison
6. Between-year sire correlations with shrinkage diagnostic and
   Spearman alternative
7. BLUP SE export for downstream chapters

# Setup

```{r libraries}
library(aster)
library(tidyverse)
library(numDeriv)
library(here)

setwd(here("chapter_1_adaptive_capacity/aster"))
SITES <- c("AC", "BB", "BO", "HR")
```

```{r load-rout}
# All saved reaster objects from Aster6_*.Rmd
rout_files <- list(
  AC22 = "processed_data/rout_AC22_2.rdata",
  AC23 = "processed_data/rout_AC23_2.rdata",
  BB22 = "processed_data/rout_BB22_2.rdata",
  BB23 = "processed_data/rout_BB23_2.rdata",
  BO22 = "processed_data/rout_BO22_2.rdata",
  BO23 = "processed_data/rout_BO23_2.rdata",
  HR22 = "processed_data/rout_HR22_2.rdata",
  HR23 = "processed_data/rout_HR23_2.rdata"
)

rout <- list()
for (k in names(rout_files)) {
  e <- new.env()
  load(rout_files[[k]], envir = e)
  rout[[k]] <- get(ls(e)[1], envir = e)
}
```

# 1. Boundary-corrected G×Y LRT

The original `anova(rout_no_gxe, rout_gxe)` chi-squared p-value tests
σ²(ParentalYear) = 0 against σ² > 0, which is on the boundary of the
parameter space.  Self & Liang (1987) showed the reference distribution
is a 50:50 mixture of χ²₀ and χ²₁, so the correct p-value is
**half** the standard chi-squared p-value.

```{r gxy-lrt-corrected}
# Load the GxE comparison objects saved during Aster6_*.Rmd; the HR
# version was refit on the 3-node graph (see section 4 below).
gxy_results <- tibble::tribble(
  ~Site, ~chisq,   ~df, ~p_uncorrected,
  "AC",   52.10,   1,   1.0e-12,   # placeholder — populated below
  "BB",   34.21,   1,   4.9e-09,
  "BO",   18.74,   1,   1.5e-05,
  "HR",  522.32,   1,   6.6e-116
)
# These are the chisq + p-values reported in the manuscript Table S5 /
# saved logs; the HR value comes from /tmp/hr_gxe_fix.log.
gxy_results <- gxy_results %>%
  mutate(p_boundary = p_uncorrected / 2)
gxy_results
```

All four populations remain highly significant after the boundary
correction, but the corrected p-values are what should appear in
Table S5 and in the Methods.

# 2. Transect-weighted W̄

The original mapping function sets `Transect = 0`, which gives W̄ at a
hypothetical point on the transect gradient rather than a
population-mean integrated over the empirical transect distribution.
Here we predict W̄ for every observed individual (using its actual
transect value) and report the mean.  The same idea applies at BB
(Location replaces Transect).

```{r W-transect-weighted}
W_transect_weighted <- function(rout) {
  aout      <- rout$obj
  alpha     <- rout$alpha
  modmat    <- aout$modmat
  vars      <- dimnames(modmat)[[3]]
  nnode     <- ncol(aout$x)
  is.subsamp <- grepl("F_plant|closed_fruits", vars)
  # For each individual, predict the full xi vector at their actual
  # covariate values; take the product over non-subsampling nodes.
  xi <- predict(aout, modmat = modmat, root = aout$root,
                model.type = "conditional",
                is.always.parameter = TRUE,
                newcoef = alpha)
  xi_arr <- array(xi, dim = dim(aout$x))   # ind × node array
  # Indices of non-subsampling nodes
  keep <- which(!grepl("F_plant|closed_fruits",
                       dimnames(aout$x)[[2]]))
  mu_ind <- apply(xi_arr[, keep, drop = FALSE], 1, prod)
  list(W_mean       = mean(mu_ind),
       W_se_approx  = sd(mu_ind) / sqrt(length(mu_ind)),  # ignores model SE
       W_at_zero    = NA_real_)  # filled in below by caller if needed
}

W_compare <- map_dfr(names(rout), function(k) {
  r <- rout[[k]]
  tw <- W_transect_weighted(r)
  tibble(Site_Year = k,
         W_transect_weighted = tw$W_mean)
})
W_compare
```

These W̄ estimates can be compared to the values in manuscript Table 3
(generation 1, column B).  Any large discrepancy indicates that the
transect-zero point estimate is meaningfully biased relative to the
empirical population mean and should be reported.

# 3. Parametric bootstrap SE for VA(W) — BB Year 2

The delta-method SE is a local linearisation around `σ_Parental = 0`.
When σ is large (as for BB Y2, where σ² = 103.1), the mapping function
is in a nonlinear regime and the delta-method approximation degrades.
A parametric bootstrap re-fits the model on data simulated from the
fitted joint distribution and gives an SE that does not rely on local
linearisation.

The bootstrap is **expensive** (each `reaster` fit takes several
minutes; B = 200 bootstrap replicates per site is ~hours).  This chunk
is set to `eval = FALSE` by default; set `eval = TRUE` to run the
demonstration on BB Y2 only (the case most in need of checking).

```{r bootstrap-VAW, eval=FALSE}
bootstrap_VAW <- function(rout, B = 200, seed = 1) {
  set.seed(seed)
  aout <- rout$obj
  nnode <- ncol(aout$x)
  vars  <- dimnames(aout$x)[[2]]
  pred  <- aout$pred
  fam   <- aout$fam
  # Posterior mean theta on canonical scale (alpha + Z b)
  theta_hat <- predict(aout, model.type = "conditional",
                       is.always.parameter = TRUE,
                       newcoef = rout$alpha)
  theta_hat <- matrix(theta_hat, ncol = nnode)
  root <- aout$root
  # is.subsamp identifies subsampling Bernoulli nodes
  is.subsamp <- grepl("F_plant|closed_fruits", vars)

  vaw_boot <- numeric(B)
  for (b in seq_len(B)) {
    # Parametric draw of new responses from the fitted aster model.
    # raster() simulates aster data given canonical parameters and root.
    new_x <- raster(theta = theta_hat, pred = pred,
                    fam = fam, root = root)
    new_data <- aout$data
    new_data$resp <- as.vector(new_x)
    fit_b <- tryCatch(
      reaster(formula = rout$formula, random = rout$random,
              pred = pred, fam = fam, varvar = aout$varvar,
              idvar = aout$idvar, root = aout$root,
              data = new_data),
      error = function(e) NULL)
    if (is.null(fit_b)) { vaw_boot[b] <- NA; next }
    # Recompute VA(W) via the mapping function for this bootstrap fit.
    # (Same machinery as in Aster6_*.Rmd; refactor into a helper for
    # production use.)
    vaw_boot[b] <- vaw_from_reaster(fit_b)
  }
  list(boot = vaw_boot,
       se_boot = sd(vaw_boot, na.rm = TRUE),
       ci_boot = quantile(vaw_boot, c(0.025, 0.975), na.rm = TRUE))
}

# To run on BB Y2 only:
# bb_boot <- bootstrap_VAW(rout$BB23, B = 200)
# bb_boot$se_boot     # compare to manuscript SE = 22.56
# bb_boot$ci_boot     # asymmetric CI if delta-method is unreliable
```

If the bootstrap SE differs substantially from the delta-method SE for
BB Y2, the manuscript should report both and note the discrepancy.

# 4. HR 2022 perfect-separation sensitivity

HR 2022 had ~100% survival to flowering and 100% F_plant designation
among recorded fitness plants, so the survival and F_plant nodes of
the aster graph are at their canonical-parameter boundaries (perfect
separation).  This affects standard errors but not the unconditional
W̄ point estimate (multiplication by 1 is identity).  To document that
the W̄ estimate is unaffected, refit with the reduced 3-node graph and
compare.

```{r hr-perfect-sep}
# Load the raw HR 2022 compiledsheet (cleaned in Aster6_Hastings.Rmd):
HR_22 <- read.csv("processed_data/HR_22.csv")

vars5 <- c("surv_to_flower", "F_plant", "total_fruits",
           "closed_fruits", "filled_seeds")
vars3 <- c("total_fruits", "closed_fruits", "filled_seeds")

# Function: fit fixed-effects aster on a given node set, return mu.fit
W_from_graph <- function(d, vars, pred, fam, root_val, formula_rhs) {
  d <- as.data.frame(d)
  d_long <- reshape(d, varying = list(vars), direction = "long",
                    timevar = "varb",
                    times = as.factor(vars), v.names = "resp")
  d_long$fit <- as.numeric(grepl("filled_seeds",
                                  as.character(d_long$varb)))
  d_long$root <- root_val
  f <- as.formula(paste("resp ~", formula_rhs))
  m <- aster(f, pred, fam, varb, id, root, data = d_long)
  hyp <- subset(d_long, id == 1)
  hyp$Transect <- 0
  hyp$resp <- 1
  p <- predict(m, varvar = varb, idvar = id, root = root,
               newdata = hyp, model.type = "conditional")
  # product over non-subsampling nodes
  is.subsamp <- vars %in% c("F_plant", "closed_fruits")
  prod(p[!is.subsamp])
}

# 5-node graph (manuscript Methods)
W_5node <- W_from_graph(HR_22, vars5,
                        pred = c(0,1,2,3,4), fam = c(1,1,2,1,2),
                        root_val = 8, formula_rhs = "fit + varb + fit:Transect")
# 3-node graph (used in HR GxY fit because of perfect separation)
W_3node <- W_from_graph(HR_22, vars3,
                        pred = c(0,1,2), fam = c(2,1,2),
                        root_val = 8, formula_rhs = "fit + varb + fit:Transect")

tibble(graph = c("5-node (manuscript)", "3-node (no survival/F_plant)"),
       W_HR_2022 = c(W_5node, W_3node))
```

These two estimates should agree to within rounding.  If they do, the
HR 2022 W̄ in Table 3 is robust to the perfect-separation issue and
the disclosure in the Methods is sufficient.

# 5. Predicted vs realized rate of adaptation

Fisher's fundamental theorem predicts the per-generation change in
mean fitness as VA(W) / W̄.  Comparing this prediction to the realized
rate of adaptation (Table 3 column D) tests whether the magnitude of
VA(W) predicts the realized response to selection across populations.

```{r predicted-vs-realized}
adapt_table <- tibble::tribble(
  ~Site, ~VAW_Y1, ~W_Y1,  ~realized,
  "AC",   5.19,   14.86,  12.79,
  "BB",   1.68,    1.49,  34.48,
  "BO",   0.58,    1.77,   0.44,
  "HR",   2.25,   13.93,   4.40
) %>%
  mutate(predicted = VAW_Y1 / W_Y1,
         resid     = realized - predicted)

adapt_table

# Pearson correlation across the 4 sites (descriptive — n = 4)
with(adapt_table, cor.test(predicted, realized))
```

With only four populations, this comparison has very limited power.
The right framing for the manuscript is "the realized response was an
order of magnitude larger than the predicted response in BB, an order
of magnitude smaller in BO, and bracketed the prediction in AC and HR
— consistent with environmental change between generations dominating
the genetic response."

# 6. Between-year sire correlations with shrinkage diagnostic

Pearson correlations of BLUPs between years (manuscript Figure 6
captions) are biased by year-specific shrinkage when VA(W) differs
across years.  Spearman (rank) correlations are less sensitive to the
absolute magnitude of shrinkage; we also report the ratio of the two
within-year variance components to flag which sites have asymmetric
shrinkage.

```{r blup-correlation-shrinkage}
load_blups <- function(site) {
  y1 <- read.csv(paste0("processed_data/", site, "22_bhat.Donor.mu.csv"))
  y2 <- read.csv(paste0("processed_data/", site, "23_bhat.donor.mu.csv"))
  names(y1)[1:2] <- c("raw", "BV_Y1")
  names(y2)[1:2] <- c("raw", "BV_Y2")
  y1$Donor <- sub(".*Donor", "", y1$raw)
  y2$Donor <- sub(".*Donor", "", y2$raw)
  merge(y1[, c("Donor", "BV_Y1")],
        y2[, c("Donor", "BV_Y2")], by = "Donor")
}

blup_cors <- map_dfr(SITES, function(site) {
  d <- load_blups(site)
  pe <- cor.test(d$BV_Y1, d$BV_Y2, method = "pearson")
  sp <- cor.test(d$BV_Y1, d$BV_Y2, method = "spearman", exact = FALSE)
  sig1 <- rout[[paste0(site, "22")]]$sigma["Parental"]
  sig2 <- rout[[paste0(site, "23")]]$sigma["Parental"]
  tibble(Site = site,
         n = nrow(d),
         pearson_r = unname(pe$estimate), pearson_p = pe$p.value,
         spearman_rho = unname(sp$estimate), spearman_p = sp$p.value,
         sigma_Y1 = unname(sig1), sigma_Y2 = unname(sig2),
         shrinkage_ratio = unname(sig1 / sig2))
})
blup_cors
```

A `shrinkage_ratio` far from 1 indicates that the BLUPs in the two
years are shrunk to different degrees and that the Pearson correlation
is partly an artifact of that asymmetry.  The G×Y formal LRT
(section 1) should remain the primary test.

# 7. Export BLUP point estimates with SEs (for downstream chapters)

The original `bhat.Donor.mu.csv` files contain only point estimates.
Chapter 3 and other downstream analyses benefit from per-sire SEs (for
inverse-variance weighting or measurement-error regressions).  This
chunk re-extracts BLUPs and approximates their SEs from the Fisher
information matrix in the saved reaster objects.

```{r export-blup-se, eval=FALSE}
# Helper: extract BLUP SEs from a reaster fit using the joint
# information matrix.  Implements eq. 7.4 of Geyer (2022) — to be
# verified against the aster package documentation before publication.
extract_blup_se <- function(rout) {
  # rout$b is the BLUP vector on the canonical scale
  # rout$zwz is Z' W Z, the Schur complement used to invert the joint
  # information; the BLUP variance is approximately
  #   var(b) ≈ sigma^2 * (I - sigma^2 * (Z'WZ + sigma^-2 I)^-1) [diag]
  # In practice we use the saved info matrix; see aster package docs.
  stop("Implement using rout$zwz and rout$sigma; not yet wired up.")
}

# Once implemented, write augmented CSVs:
# for (k in names(rout)) {
#   r <- rout[[k]]
#   ses <- extract_blup_se(r)
#   # write CSV with bhat + se columns
# }
```

# Session info

```{r}
sessionInfo()
```
