---
title: "Aster6_Hastings"
author: "Helen Payne"
date: "2025-07-30"
output:
  pdf_document: default
  html_document: default
---

# Hastings Ranch Generation 1 2022
```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
library(here)

#Read in the data
HR_22<- read_csv(here::here("data_sheets/compiled_sheets/HR_compiledsheet_full_2022.csv")) 

#organize the data
HR_22 <- HR_22 %>%
  mutate(
    sample_ID = str_c(Year, Recipient, Gen, str_pad(Transect, width = 3, pad = "0"), Plant_ID, sep = "-"), #create an id column for each individual
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"), #use "-" rather than "_"
    F_plant = ifelse(F_plant == TRUE, 1, 0), #make fitness plant binary (1/0) rather than true/false
    Transect = as.numeric(Transect)) #make sure location is a continuous variable

#clean the data
HR_22 <- HR_22 %>%
  mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower), #If a plant produced any fruits, it must have survived to flowering, even if the original surv_to_flower entry was coded as 0, This corrects miscodings where survival to flowering was mistakenly marked as 0 despite having fruit production
  ) %>%
  dplyr::select("Transect", "Donor", "Recipient", "sample_ID", "surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds") %>% #select the columns we want
  filter(!is.na(Transect))# We are removing W1 and W2 from these analyses because they were part of a different experiment involving a watered treatment. 
## In the original dataset, we included some extra fitness plants (see `anyFit_P`) in other transects for comparison with W1 and W2. 
### For the purposes of these analyses, we will exclude them and focus only on the `F_plant` values that are recorded for every segment.

HR_22$F_plant[is.na(HR_22$F_plant)] <- 0 #we need to make F_plant 0 and not NAs

#Now we want to make sure every segment is represented by one F_plant (because we will filter by F_plant soon)
HR_22 <- HR_22 %>%
  mutate(sample_prefix = substr(sample_ID, 1, 19)) %>% #create a sample prefix with the first 19 characters of sample ID
  group_by(sample_prefix) %>% #group by the prefix
  mutate(
    all_dead = all(surv_to_flower == 0), #flags groups where no plants survived and where there are currently no fitness plants
##if a group has no fitness plants, it randomly assigns one plant in the group to be the fitness plant
    no_f_plant = sum(F_plant) == 0, 
    F_plant = if_else(
      no_f_plant & row_number() == sample(n(), 1),
      1L,
      F_plant
    ) #this will pick different individuals within the group each time unless you set a seed. For the purposes of this analyses, it will not matter because their values will be zero either way.
  ) %>%
  ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant) #ungroup and remove the rows that were created\


#Now we want to determine how many individuals within a segment survived to flower
HR_22 <- HR_22 %>%
  mutate(group = ceiling(row_number() / 8)) %>%  # Create a grouping variable for every 8 rows
  # Calculate the sum of 'surv_to_flower' within each group
  group_by(group) %>%
  mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>% # Ungroup to avoid grouped data frame
  # Remove the temporary 'group' column
  dplyr::select(-group)


#Filter to just fitness plants
HR_22_filtered <- HR_22 %>%
  filter(F_plant == 1)

#Reorder for compatibility
HR_22 <- HR_22_filtered %>%
  dplyr::select(Transect, Donor, Recipient, sample_ID, total_surv_to_flower, F_plant, total_fruits, closed_fruits, filled_seeds)

#Rename for compatibility
HR_22 <- HR_22 %>%
  rename(surv_to_flower = total_surv_to_flower)

#change F_plant to 0 if survival is 0
##this will allow aster to function- F_plant cannot be 1 if survival is 0. 
HR_22 <- HR_22 %>%
  mutate(F_plant = ifelse(surv_to_flower == 0, 0, F_plant))
```


```{r}
#check for nonsense data
subset(HR_22, filled_seeds > 0 & surv_to_flower == 0)
subset(HR_22, filled_seeds > 0 & total_fruits == 0)
subset(HR_22, filled_seeds > 0 & closed_fruits == 0)

#Make  Donors and Recipients factors
HR_22$Donor <- as.factor(HR_22$Donor)
HR_22$Recipient <- as.factor(HR_22$Recipient)

#number that germinated from table 1
HR_22 %>%
  mutate(surv_to_flower = as.factor(surv_to_flower)) %>% 
  group_by(surv_to_flower) %>% 
  count()

#distinct number of Recipients
HR_22 %>%
  distinct(Recipient) %>%
  count()
#137

#distinct number of Donors
HR_22  %>%
  distinct(Donor) %>%
  count()
#46

#number of sires, and number of individuals for each recipient
NDS<-HR_22 %>%
  group_by(Recipient, Donor) %>% 
  count()

```

```{r}
#make NAs for other life history stages zero
HR_22$surv_to_flower[is.na(HR_22$surv_to_flower)] <- 0
HR_22$total_fruits[is.na(HR_22$total_fruits)]<-0
HR_22$closed_fruits[is.na(HR_22$closed_fruits)]<-0
HR_22$filled_seeds[is.na(HR_22$filled_seeds)]<-0

# Define the update_zeros function modifies the data and fixes cases where there are any inconsistencies; because the data looks good, it shouldn't change anything for this df
update_zeros <- function(df) {
  for (i in 1:nrow(df)) {
    if (df$surv_to_flower[i] == 0) {
      df$total_fruits[i] <- 0
      df$closed_fruits[i] <- 0
      df$filled_seeds[i] <- 0
    }
    if (df$total_fruits[i] == 0) {
      df$closed_fruits[i] <- 0
      df$filled_seeds[i] <- 0
    }
    if (df$closed_fruits[i] == 0) {
      df$filled_seeds[i] <- 0
    }}
  return(df)
  }

# Apply the function to the dataset
HR_22 <- update_zeros(HR_22)

write.csv(HR_22, "processed_data/HR_22.csv", row.names = FALSE)
```

```{r}
#Check the distribution of zeros in your model

# Determine how many rows are equal to 0 for the column "surv_to_flower"
surv_to_flower_zero <- sum(HR_22$surv_to_flower == 0)
cat("Number of rows where surv_to_flower is 0:", surv_to_flower_zero, "\n")
#2

# Determine how many rows are equal to 0 for the column "total_fruits"
total_fruits_zero <- sum(HR_22$total_fruits == 0)
cat("Number of rows where total_fruits is 0:", total_fruits_zero, "\n")
#36

# Determine how many rows are equal to 0 for the column "closed_fruits"
closed_fruits_zero <- sum(HR_22$closed_fruits == 0)
cat("Number of rows where closed_fruits is 0:", closed_fruits_zero, "\n")
#39

# Determine how many rows are equal to 0 for the column "filled_seeds"
filled_seeds_zero <- sum(HR_22$filled_seeds == 0)
cat("Number of rows where filled_seeds is 0:", filled_seeds_zero, "\n")
#67
```

## Set up aster
```{r}
library(aster)
library(reshape2)

#the "vars" are the nodes in the aster graphical model.   
vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

#convert to a dataframe for reshape to work
HR_22 <- as.data.frame(HR_22)

#Reshape HR to longform. "varying" notes the set of variables that will be converted to longform and named in "timevar" under one column of varb. "Times" says to use vars for the newly created column varb 
HR22_aster <- reshape(HR_22 ,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#####Check that the reshape worked 
#If it did the number of rows will be the same 
nrow(HR22_aster)
nrow(HR_22)*length(vars)

#####Designate fitness variable, "filled_seeds", which is the final node 
#grep1 is like a search function
fit<-grepl("filled_seeds", as.character(HR22_aster$varb))
fit<-as.numeric(fit)
HR22_aster$fit <- fit

#####Check "filled_seeds", is designated as the fitness variable 
with(HR22_aster, sort(unique(as.character(varb)[fit==0])))
with(HR22_aster, sort(unique(as.character(varb)[fit==1])))

#####Add "root" to HR22_aster files where value is 1 
HR22_aster <- data.frame(HR22_aster, root=8)

#####Set graphical node and distribution for fitness nodes(preds) 
names(HR22_aster)

#pred is nodes. There are 5 nodes, so five preds.
pred <- c(0,1,2,3,4)

#fam assigns a distribution to each node. 1= bernoulli, 2= poission. The subsampling nodes (node #2 and #4 in pred) are binomial so Bernouli (always). Bernouli is a specific type of binomial so the assignment for family distribution is the same
fam<-c(1,1,2,1,2)
```

## Random Effects Aster Model
```{r}
####Testing for Donor and Recipient significance
#Donor is zero with Recipient
Both_HR22 <- reaster(resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 + 
    fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = HR22_aster)
summary(Both_HR22)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 + 
#     fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, 
#     fam = fam, varvar = varb, idvar = id, root = root, data = HR22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)          1.580187   0.046741  33.808  < 2e-16 ***
# fit                 -1.271917   0.053543 -23.755  < 2e-16 ***
# varbF_plant        -14.612118   0.187750 -77.827  < 2e-16 ***
# varbsurv_to_flower   0.828214   0.069803  11.865  < 2e-16 ***
# varbtotal_fruits    -0.488937   0.084326  -5.798  6.7e-09 ***
# fit:Transect         0.004499   0.001279   3.517 0.000437 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#            Estimate Std. Error z value Pr(>|z|)/2    
# Donor     0.0001838  0.6003500   0.000        0.5    
# Recipient 0.0312676  0.0051740   6.043   7.55e-10 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


#Check for the effect of Donor alone. Significant. 
Donor_HR22<- reaster(resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 + 
    fit:Donor), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = HR22_aster)
summary(Donor_HR22)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 + 
#     fit:Donor), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = HR22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)          1.579807   0.046742  33.798  < 2e-16 ***
# fit                 -1.266350   0.053505 -23.668  < 2e-16 ***
# varbF_plant        -14.648654   0.187615 -78.078  < 2e-16 ***
# varbsurv_to_flower   0.828594   0.069804  11.870  < 2e-16 ***
# varbtotal_fruits    -0.488557   0.084326  -5.794 6.89e-09 ***
# fit:Transect         0.004336   0.001258   3.446 0.000569 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#       Estimate Std. Error z value Pr(>|z|)/2    
# Donor 0.019605   0.004602    4.26   1.02e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Recipient_HR22 <- reaster(resp ~ fit + varb + fit:Transect, random = list(Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = HR22_aster)
summary(Recipient_HR22)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Transect, random = list(Recipient = ~0 + 
#     fit:Recipient), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = HR22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)          1.580187   0.046741  33.807  < 2e-16 ***
# fit                 -1.271917   0.053543 -23.755  < 2e-16 ***
# varbF_plant        -14.612112   0.187749 -77.828  < 2e-16 ***
# varbsurv_to_flower   0.828214   0.069803  11.865  < 2e-16 ***
# varbtotal_fruits    -0.488937   0.084326  -5.798  6.7e-09 ***
# fit:Transect         0.004499   0.001279   3.518 0.000435 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#           Estimate Std. Error z value Pr(>|z|)/2    
# Recipient 0.031269   0.004135   7.561      2e-14 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


####Combine Sire and Dam into one random effect, `Parental`. The estimates are of similar magnitude 
modmat.Donor <- model.matrix(~ 0 + fit:Donor, HR22_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, HR22_aster)
modmat.donorrecipient <- cbind(modmat.Donor,modmat.Recipient)

rout_HR22 <- reaster(resp ~ fit + varb, list(Parental = ~ 0 + modmat.donorrecipient),pred, fam, varb, id, root, data = HR22_aster)
sout_HR22 <-summary(rout_HR22)
sout_HR22

#Call:
# reaster.formula(fixed = resp ~ fit + varb, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = HR22_aster)
# 
# 
# Fixed Effects:
#                     Estimate Std. Error z value Pr(>|z|)    
# (Intercept)          1.57999    0.04674  33.803  < 2e-16 ***
# fit                 -1.24691    0.05308 -23.492  < 2e-16 ***
# varbF_plant        -14.63153    0.18779 -77.916  < 2e-16 ***
# varbsurv_to_flower   0.82841    0.06980  11.868  < 2e-16 ***
# varbtotal_fruits    -0.48874    0.08433  -5.796  6.8e-09 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.022409   0.003409   6.574   2.45e-11 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#model comparison for significance of Transect. Significant.
rout_HR22_2 <- reaster(resp ~ fit + varb + fit:Transect, list(Parental = ~ 0 + modmat.donorrecipient),
    pred, fam, varb, id, root, data = HR22_aster)
sout_HR22_2 <-summary(rout_HR22_2)
sout_HR22_2

save(rout_HR22_2, file = "processed_data/rout_HR22_2.rdata")

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Transect, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = HR22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)          1.580119   0.046741  33.806  < 2e-16 ***
# fit                 -1.271109   0.053609 -23.711  < 2e-16 ***
# varbF_plant        -14.616677   0.187783 -77.838  < 2e-16 ***
# varbsurv_to_flower   0.828282   0.069803  11.866  < 2e-16 ***
# varbtotal_fruits    -0.488869   0.084326  -5.797 6.74e-09 ***
# fit:Transect         0.004422   0.001272   3.475  0.00051 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.022499   0.003429   6.562   2.66e-11 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova(rout_HR22, rout_HR22_2)

#Analysis of Deviance Table
# 
# Model 1: resp ~ fit + varb, ~0 + modmat.donorrecipient
# Model 2: resp ~ fit + varb + fit:Transect, ~0 + modmat.donorrecipient
#   Mod Df Fix Mod Df Rand Mod Dev Df Fix Df Rand Deviance    P-value
# 1          5           1 -1605.9                                   
# 2          6           1 -1593.5      1       0   12.439 0.00042043

####Now look at parameter estimates to be sure we have the model intended.
sout_HR22_2 <-summary(rout_HR22_2)
sout_HR22_2
```

## Mapping Function

```{r}
#following Geyer et al. 2022
library(numDeriv)
fishinv<-solve(sout_HR22_2$fisher)
map.factory <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  modmat <- rout$obj$modmat[1, , , drop = FALSE]
  modmat[ , , "fit:Transect"] <- 0
  # set root = 8
  root <- array(8, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, 5 is the number of nodes 
  print(x)
  # return map function
  function (b) {
    stopifnot(is.numeric(b))
    stopifnot(is.finite(b))
    stopifnot(length(b) == 1)
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, # added
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      is.always.parameter = TRUE)
  
    xi <- matrix(xi, ncol = nnode)
    # always use drop = FALSE unless you are sure you don't want that
    # here if we omit drop = FALSE and there is only one non-subsampling
    # node, the code will break (apply will give an error)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod) #every individual
    # mu is unconditional mean values for model without subsampling
    # in this application all components mu are the same because no
    # covariates except varb, so just return only one
    mu[1] 
  }
}

map <- map.factory(rout_HR22_2, vars %in% c("F_plant", "closed_fruits"))


#MEAN FITNESS
mf<-map(0)
mf
#13.92531

#ADDITIVE GENETIC VARIANCE FOR FITNESS
vaw <- rout_HR22_2$sigma["Parental"]^2 * grad(map, 0)^2
# get rid of name
vaw <- as.numeric(vaw)
vaw <- vaw * 4
vaw
#2.250045

####FFTNS
fftns <- vaw / mf
fftns
#0.1615795

#prediction of progeny generation mean fitness
mf+fftns
#14.08689

# get SE
map.factory.too <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  # modmat for one individual. Taking the first row of modmat- there is one row for each individual
  modmat <- rout$obj$modmat[1, , , drop = FALSE] #this only takes the first row 
  modmat[ , , "fit:Transect"] <- 0
  # set root = 8
  root <- array(8, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, second number is number of nodes 
  # return map function
  function (balpha) {
    stopifnot(is.numeric(balpha))
    stopifnot(is.finite(balpha))
    stopifnot(length(balpha) == 1 + length(alpha))
    b <- balpha[1]
    alpha <- balpha[-1]
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, # added
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      is.always.parameter = TRUE
    )
    xi <- matrix(xi, ncol = nnode)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod)
    # mu is unconditional mean values for model without subsampling
    mu[1]
  }
}

map.too <- map.factory.too(rout_HR22_2, vars %in% c("F_plant", "closed_fruits"))


#centering sire on zero
balpha.hat <- c(0, rout_HR22_2$alpha)

map.too(balpha.hat)
#13.92531

all.equal(map(0), map.too(balpha.hat))

#first and second derivatives
g<-grad(map.too, balpha.hat)

#calculates the nxn matrix
h <- hessian(map.too, balpha.hat)

#these are partial derivatives
dmu.db <- g[1]
dmu.dalpha <- g[-1]
d2mu.db.dalpha <- h[1, -1]

# give names to the estimators in our formulas
mu.hat <- map.too(balpha.hat)
nu.hat <- rout_HR22_2$nu["Parental"]

#FFTNS SE
#calculations for the gradient vector of the FFTNS prediction with respect to the parameters of the models (fixed effects and variance components) 
#zero removed
dfftns <- c(- 4 * nu.hat * dmu.dalpha * dmu.db^2 / mu.hat^2 +
8 * nu.hat * d2mu.db.dalpha / mu.hat, 4 * dmu.db^2 / mu.hat)

#And apply the delta method.
fftns.se <- t(dfftns) %*% fishinv %*% dfftns
fftns.se <- sqrt(as.vector(fftns.se))
fftns.se
#0.05009587

#ADDITIVE GENETIC VARIANCE SE 
#calculate the gradient vector VA(W) with respect to the parameters of the model using these formulae
dvaw <- c(8 * nu.hat * d2mu.db.dalpha, 4 * dmu.db^2)

#And then apply the delta method to get standard errors for this estimator
vaw.se <- t(dvaw) %*% fishinv %*% dvaw
vaw.se <- sqrt(as.vector(vaw.se))
vaw.se
#0.6848743

#MEAN FITNESS SE
dmf <- c(dmu.dalpha, 0)
mf.se <- t(dmf) %*% fishinv %*% dmf
mf.se <- sqrt(as.vector(mf.se))
mf.se
#0.3976455
```

## Breeding Values

```{r}
#extract breeding values
bhat <- rout_HR22_2$b 
bhat.Donor<- bhat[grep("Donor", names(bhat))]

#check this is the correct number of sires (46)
length(bhat.Donor)

#convert canonical values to mean value parameter values using mapping function.
map <- map.factory(rout_HR22_2, vars %in% c("F_plant", "closed_fruits"))
vectorized.map <- Vectorize(map)
curve(vectorized.map, from = -1 / 2, to = 1/2, xlab="b", ylab=expression(mu(b)))

#breeding values for sire groups on the mean value parameter scale 
bhat.Donor.mu <- vectorized.map(bhat.Donor)
bhat.Donor.mu<-as.data.frame(bhat.Donor.mu)
dmap_HR22 <- Vectorize(function(b) numDeriv::grad(map, b))
bhat.Donor.mu$BV_se_approx <- abs(dmap_HR22(bhat.Donor)) *
  rout_HR22_2$sigma["Parental"]
write.csv(bhat.Donor.mu, "processed_data/HR22_bhat.Donor.mu.csv")

#plot the distribution of breeding values
# Extract the numeric column from bhat.Donor.mu
numeric_values <- bhat.Donor.mu[["bhat.Donor.mu"]]  # Column name is the same as the data frame name

# Calculate and plot the density
prob_den <- density(numeric_values)
plot(prob_den)
```
```{r}
#BACK OF NAPKIN FITNESS CALCULATION

W <- HR_22 %>%
  summarise(
    total_seeds = sum(filled_seeds, na.rm = TRUE),
    W = total_seeds / n()
  )

W 
#15.13625	
```


# Hastings Ranch Generation 1 2023
```{r}
#load year 2 (2023) life history data for HR site
HR_23 <- read_csv(here::here("data_sheets/compiled_sheets/HR_compiledsheet_full_2023.csv")) 

#organize the data
HR_23 <- HR_23 %>%
  dplyr::select("Transect", "Gen", "Donor", "Recipient", "sample_ID_SEG", "surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds", "msm_all") %>%
  mutate(F_plant = ifelse(F_plant == TRUE, 1, 0)) %>%
  mutate(surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower)) %>% #if a plant made fruits, it must have survived to flowering, so the survival indicator is changed from 0 to 1 
  filter(Donor != "random seeds")

HR_23 <- HR_23 %>%
  # Create a grouping variable for every 5 rows
  mutate(group = sample_ID_SEG) %>%
  # Calculate the sum of 'surv_to_flower' within each group
  group_by(group) %>%
  mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>% # Ungroup to avoid grouped data frame
  # Remove the temporary 'group' column using dplyr::select
  dplyr::select(-group)

#Ensure at least one F_plant per sample_ID prefix (first 19 characters)
HR_23 <- HR_23 %>%
  group_by(sample_ID_SEG) %>%
  arrange(Transect, Recipient, Donor, sample_ID_SEG, .by_group = TRUE) %>%
  mutate(
    has_f = any(F_plant == 1),
    m_surv = sum(surv_to_flower == 1, na.rm = TRUE),
    # if none marked, pick deterministically:
    F_plant = case_when(
      has_f ~ F_plant,# keep existing
      !has_f & m_surv == 0 & row_number() == 1L ~ 1L,    # no survivors: pick first row (placeholder)
      TRUE ~ F_plant
    )
  ) %>%
  ungroup() %>%
  dplyr::select(-has_f, -m_surv)


#Filter to just fitness plants
HR_23 <- HR_23 %>%
  filter(F_plant == 1)

HR_23_rearranged <- HR_23 %>%
  dplyr::select(Transect, Gen, Donor, Recipient, sample_ID_SEG, total_surv_to_flower, F_plant, total_fruits, closed_fruits, filled_seeds, msm_all)


HR_23_rearranged <- HR_23_rearranged %>%
  rename(surv_to_flower = total_surv_to_flower)


#change F_plant to 0 if survival is 0 to let aster work
HR_23 <- HR_23_rearranged %>%
  mutate(F_plant = ifelse(surv_to_flower == 0, 0, F_plant))
```



```{r}
####make NAs for other life history stages zero
HR_23[is.na(HR_23$surv_to_flower),]$surv_to_flower <- 0
HR_23[is.na(HR_23$total_fruits),]$total_fruits <- 0
HR_23[is.na(HR_23$closed_fruits),]$closed_fruits <- 0
HR_23[is.na(HR_23$filled_seeds),]$filled_seeds <- 0


####check for nonsense data
subset(HR_23, filled_seeds > 0 & surv_to_flower  == 0)
subset(HR_23, filled_seeds > 0 & total_fruits == 0)
subset(HR_23, filled_seeds > 0 & closed_fruits ==0)
subset(HR_23, filled_seeds > 0 & filled_seeds ==0)
subset(HR_23, closed_fruits > total_fruits)
subset(HR_23, total_fruits > 0 & surv_to_flower == 0)


#make factors
HR_23$Donor <- as.factor(HR_23$Donor)
HR_23$Recipient <- as.factor(HR_23$Recipient)
HR_23$Gen<-as.factor(HR_23$Gen)

```

## Germination in G1 vs G2 (parental vs progeny)
```{r}
library(tidyverse)
#data for table 1
#counts the number of Recipients in parental and progeny generation
HR_23 %>%
  group_by(Gen) %>% 
  distinct(Recipient) %>%
  arrange(Gen, Recipient) %>%
  count()

#counts the number of Donors in each group. For progeny number of sires is grand-Donors as they were open pollinated in the field 
HR_23 %>%
  group_by(Gen) %>% 
  distinct(Donor) %>%
  count()

HR_23 %>%
  group_by(Gen, Recipient, Donor) %>% 
  count()
```

```{r}
#Check the distribution of zeros in your model

# Determine how many rows are equal to 0 for the column "surv_to_flower"
surv_to_flower_zero <- sum(HR_23$surv_to_flower == 0)
cat("Number of rows where surv_to_flower is 0:", surv_to_flower_zero, "\n")
#32

# Determine how many rows are equal to 0 for the column "total_fruits"
total_fruits_zero <- sum(HR_23$total_fruits == 0)
cat("Number of rows where total_fruits is 0:", total_fruits_zero, "\n")
#135

# Determine how many rows are equal to 0 for the column "closed_fruits"
closed_fruits_zero <- sum(HR_23$closed_fruits == 0)
cat("Number of rows where closed_fruits is 0:", closed_fruits_zero, "\n")
#228

# Determine how many rows are equal to 0 for the column "filled_seeds"
filled_seeds_zero <- sum(HR_23$filled_seeds == 0)
cat("Number of rows where filled_seeds is 0:", filled_seeds_zero, "\n")
#307

```

## Seed Weight
```{r}
#regress fitness on seed weight
library(dplyr)
library(ggplot2)

seedWgt_HR <- HR_23 %>%
  dplyr::select(sample_ID_SEG, Recipient, Gen, surv_to_flower, msm_all, filled_seeds) %>%
  filter(!is.na(msm_all)) %>%
  group_by(Recipient)

# t-test of seed weight between generations
t.test(subset(seedWgt_HR, Gen == "G1")$msm_all, subset(seedWgt_HR, Gen == "G2")$msm_all)

#	Welch Two Sample t-test

# data:  subset(seedWgt_HR, Gen == "G1")$msm_all and subset(seedWgt_HR, Gen == "G2")$msm_all
# t = 0.040208, df = 453.72, p-value = 0.9679
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -0.2417400  0.2518386
# sample estimates:
# mean of x mean of y 
#  2.289560  2.284511 

#average seed weight of progeny and parental
seedWgt_HR %>%
  group_by(Gen) %>%
  summarise(Average=mean(msm_all))
```


## Set Up Aster
```{r}
#subset the data to only Gen 1
HR23_aster_a<-subset(HR_23, Gen=="G1")

HR23_aster_a <- HR23_aster_a[, !names(HR23_aster_a) %in% "msm_all"]

vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

HR23_aster_a <- as.data.frame(HR23_aster_a)

#reshape HR23_aster_a to longform
HR23_aster <- reshape(HR23_aster_a,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#Check that the reshape worked 
nrow(HR23_aster)
nrow(HR23_aster_a)*length(vars)

#designate fitness variable as terminal node, "filled_seeds"
fit<-grepl("filled_seeds", as.character(HR23_aster$varb))
fit<-as.numeric(fit)
HR23_aster$fit<-fit

#check that "filled_seeds" is designated as the fitness variable
with(HR23_aster, sort(unique(as.character(varb)[fit==0])))
with(HR23_aster, sort(unique(as.character(varb)[fit==1])))

#Add a variable "root" to HR23_aster where value is 1
HR23_aster <- data.frame(HR23_aster, root=5)

#give each node a number
pred <- c(0,1,2,3,4)

#designate each node with a distribution. 1= Bernoulli, 2= poisson. sub sampling node (pods_collected) is binomial(Bernoulli is a type of binomial distribution so the family assignment is the same)
fam<-c(1,1,2,1,2)
```

## Random Effects Aster Model
```{r}
####Fit random effects model. sire estimate is zero with dam
Both_HR23<- reaster(resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 +
    fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb,
    idvar = id, root = root, data = HR23_aster)
summary(Both_HR23)

#

#Check for the effect of sire alone. Significant. 
donor_HR23<- reaster(resp ~ fit + varb + fit:Transect, random = list(Donor = ~0 + 
    fit:Donor), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = HR23_aster)
summary(donor_HR23)

#

#check for the effect of dam alone. significant
recipient_HR23<- reaster(resp ~ fit + varb + fit:Transect, random = list(Recipient = ~0 + 
    fit:Recipient), pred = pred, fam = fam, varvar = varb, #in model dam_ML19 shouldn't this say "fit:Dam"
    idvar = id, root = root, data = HR23_aster)
summary(recipient_HR23)

#

####Combine Donor and Recipient into one random effect, `Parental`. The estimates are of similar magnitude 
modmat.Donor <- model.matrix(~ 0 + fit:Donor, HR23_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, HR23_aster)
modmat.donorrecipient <- cbind(modmat.Donor,modmat.Recipient)


rout_HR23 <- reaster(resp ~ fit + varb, list(Parental = ~ 0 + modmat.donorrecipient),pred, fam, varb, id, root, data = HR23_aster)
sout_HR23 <-summary(rout_HR23)
sout_HR23

#Call:
# reaster.formula(fixed = resp ~ fit + varb, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = HR23_aster)
# 
# 
# Fixed Effects:
#                    Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -0.24962    0.06807  -3.667 0.000245 ***
# fit                 1.03393    0.08181  12.638  < 2e-16 ***
# varbF_plant        -3.98024    0.14084 -28.260  < 2e-16 ***
# varbsurv_to_flower  1.05942    0.08393  12.623  < 2e-16 ***
# varbtotal_fruits    1.13572    0.10477  10.840  < 2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.058167   0.006894   8.438     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#model comparison for significance of yloc. Significant.
rout_HR23_2 <- reaster(resp ~ fit + varb + fit:Transect, list(Parental = ~ 0 + modmat.donorrecipient),
    pred, fam, varb, id, root, data = HR23_aster)
sout_HR23_2 <-summary(rout_HR23_2)
sout_HR23_2

save(rout_HR23_2, file = "processed_data/rout_HR23_2.rdata")

# Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Transect, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = HR23_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -0.2496338  0.0680714  -3.667 0.000245 ***
# fit                 1.0319696  0.0828051  12.463  < 2e-16 ***
# varbF_plant        -3.9803824  0.1408457 -28.261  < 2e-16 ***
# varbsurv_to_flower  1.0594296  0.0839303  12.623  < 2e-16 ***
# varbtotal_fruits    1.1357297  0.1047674  10.840  < 2e-16 ***
# fit:Transect        0.0004041  0.0026047   0.155 0.876703    
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.058031   0.006936   8.367     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova(rout_HR23, rout_HR23_2)

# Analysis of Deviance Table
# 
# Model 1: resp ~ fit + varb, ~0 + modmat.donorrecipient
# Model 2: resp ~ fit + varb + fit:Transect, ~0 + modmat.donorrecipient
#   Mod Df Fix Mod Df Rand Mod Dev Df Fix Df Rand Deviance P-value
# 1          5           1 -5556.3                                
# 2          6           1 -5556.3      1       0 0.010369 0.91889

```

## Mapping Function

```{r}
#following Geyer et al. 2022

library(numDeriv)
fishinv<-solve(sout_HR23_2$fisher)
map.factory <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  modmat <- rout$obj$modmat[1, , , drop = FALSE]
  modmat[ , , "fit:Transect"] <- 0
  # set root = 5
  root <- array(5, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, 5 is the number of nodes 
  # return map function
  function (b) {
    stopifnot(is.numeric(b))
    stopifnot(is.finite(b))
    stopifnot(length(b) == 1)
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, # added
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      is.always.parameter = TRUE)
  
    xi <- matrix(xi, ncol = nnode)
    # always use drop = FALSE unless you are sure you don't want that
    # here if we omit drop = FALSE and there is only one non-subsampling
    # node, the code will break (apply will give an error)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod) #every individual
    # mu is unconditional mean values for model without subsampling
    # in this application all components mu are the same because no
    # covariates except varb, so just return only one
    mu[1] 
  }
}

map <- map.factory(rout_HR23_2, vars %in% c("F_plant", "closed_fruits"))

#MEAN FITNESS
mf<-map(0)
mf
#5.755667

#VA(W)
vaw <- rout_HR23_2$sigma["Parental"]^2 * grad(map, 0)^2
# get rid of name
vaw <- as.numeric(vaw)
vaw <- vaw * 4
vaw
#4.919318

####FFTNS
fftns <- vaw / mf
fftns
#0.8546912

# get SE
map.factory.too <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  # modmat for one individual. Taking the first row of modmat- there is one row for each individual
  modmat <- rout$obj$modmat[1, , , drop = FALSE] #this only takes the first row
  modmat[ , , "fit:Transect"] <- 0
  # set root = 5
  root <- array(5, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, second number is number of nodes
  # return map function
  function (balpha) {
    stopifnot(is.numeric(balpha))
    stopifnot(is.finite(balpha))
    stopifnot(length(balpha) == 1 + length(alpha))
    b <- balpha[1]
    alpha <- balpha[-1]
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, # added
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      is.always.parameter = TRUE
    )
    xi <- matrix(xi, ncol = nnode)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod)
    # mu is unconditional mean values for model without subsampling
    # Here Yloc was set to zero so there in only one value in the Xi matrix 
    mu[1]
  }
}

map.too <- map.factory.too(rout_HR23_2, vars %in% c("F_plant", "closed_fruits"))

#centering sire on zero
balpha.hat <- c(0, rout_HR23_2$alpha)

map.too(balpha.hat)
#5.755667

all.equal(map(0), map.too(balpha.hat))

#first and second derivatives
g<-grad(map.too, balpha.hat)

#calculates the nxn matrix
h <- hessian(map.too, balpha.hat)

#these are partial derivatives
dmu.db <- g[1]
dmu.dalpha <- g[-1]
d2mu.db.dalpha <- h[1, -1]

# give names to the estimators in our formulas
mu.hat <- map.too(balpha.hat)
nu.hat <- rout_HR23_2$nu["Parental"]

#FFTNS SE
#calculations for the gradient vector of the FFTNS prediction with respect to the parameters of the models (fixed effects and variance components) 
#zero removed
dfftns <- c(- 4 * nu.hat * dmu.dalpha * dmu.db^2 / mu.hat^2 +
8 * nu.hat * d2mu.db.dalpha / mu.hat, 4 * dmu.db^2 / mu.hat)

#And apply the delta method.
fftns.se <- t(dfftns) %*% fishinv %*% dfftns
fftns.se <- sqrt(as.vector(fftns.se))
fftns.se
#0.2159999

#ADDITIVE GENETIC VARIANCE SE
#calculate the gradient vector VA(W) with respect to the parameters of the model using these formulae
dvaw <- c(8 * nu.hat * d2mu.db.dalpha, 4 * dmu.db^2)

#And then apply the delta method to get standard errors for this estimator
vaw.se <- t(dvaw) %*% fishinv %*% dvaw
vaw.se <- sqrt(as.vector(vaw.se))
vaw.se
#1.169627

#MEAN FITNESS SE
dmf <- c(dmu.dalpha, 0)
mf.se <- t(dmf) %*% fishinv %*% dmf
mf.se <- sqrt(as.vector(mf.se))
mf.se
#0.382223
```

## Breeding Values

```{r}
bhat <- rout_HR23_2$b 
bhat.donor<- bhat[grep("Donor", names(bhat))]

#check this is the correct number of sires (40)
length(bhat.donor)

#convert canonical values to mean value parameter values using mapping function
map <- map.factory(rout_HR23_2, vars %in% c("F_plant", "closed_fruits"))
vectorized.map <- Vectorize(map)

#breeding values for sire groups on the mean value parameter scale 
bhat.donor.mu <- vectorized.map(bhat.donor)
bhat.donor.mu<-as.data.frame(bhat.donor.mu)
dmap_HR23 <- Vectorize(function(b) numDeriv::grad(map, b))
bhat.donor.mu$BV_se_approx <- abs(dmap_HR23(bhat.donor)) *
  rout_HR23_2$sigma["Parental"]
write.csv(bhat.donor.mu, "processed_data/HR23_bhat.Donor.mu.csv")
```

# Hastings Ranch Generation 2 2023

```{r}
#subset the dataframe for the progeny generation
HR23_aster_b<-subset(HR_23, Gen=="G2")
HR23_aster_b<-droplevels(HR23_aster_b)
levels(HR23_aster_b$Gen)
```


## Set up aster

```{r}
#establish nodes as vars
vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

HR23_aster_b <- as.data.frame(HR23_aster_b)

#reshape
HR23_aster_g2 <- reshape(HR23_aster_b,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#Check that the reshape worked 
nrow(HR23_aster_g2)
nrow(HR23_aster_b)*length(vars)

#designate fitness node, "filled_seeds"
fit<-grepl("filled_seeds", as.character(HR23_aster_g2$varb))
fit<-as.numeric(fit)
HR23_aster_g2$fit<-fit

#check "Total_seeds" is designated as the fitness variable 
with(HR23_aster_g2, sort(unique(as.character(varb)[fit==0])))
with(HR23_aster_g2, sort(unique(as.character(varb)[fit==1])))

#Add a variable "root" to HR_aster_g2 files where value is 5
HR23_aster_g2 <- data.frame(HR23_aster_g2, root=5)

#give each node a number
pred <- c(0,1,2,3,4)

#designate each node with a distribution. 1= Bernoulli, 2= poisson, subsampling node is binomial which is the same family as Bernoulli
fam<-c(1,1,2,1,2)

#show graphical model
foo <- c("root", vars)
pvars <- foo[pred +1]
bar <-data.frame(pvars, vars, sapply(fam.default(), as.character)[fam])
colnames(bar) <- c("pred", "succ", "fam")
print(bar, right = FALSE, row.names = FALSE)
```

### Fixed Effect Aster Model

```{r}
HR23_g2 <- aster(resp ~ fit + varb + fit:Transect, pred, fam, varb, id, root, data = HR23_aster_g2) 

summary(HR23_g2, info.tol = 1e-9, se.fit=TRUE)

#Call:
# aster.formula(formula = resp ~ fit + varb + fit:Transect, pred = pred,
#     fam = fam, varvar = varb, idvar = id, root = root, data = HR23_aster_g2)
#
#                     Estimate Std. Error z value Pr(>|z|)
# (Intercept)        -0.439227   0.073305  -5.992 2.08e-09 ***
# fit                 1.281775   0.088802  14.434  < 2e-16 ***
# varbF_plant        -3.242391   0.137223 -23.629  < 2e-16 ***
# varbsurv_to_flower  1.433830   0.088957  16.118  < 2e-16 ***
# varbtotal_fruits    1.282477   0.109594  11.702  < 2e-16 ***
# fit:Transect       -0.005172   0.002854  -1.812   0.0699 .
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ‘ 1
#
# Original predictor variables dropped (aliased)
#      varbfilled_seeds

HR_23_hyp <- subset(HR23_aster_g2, id == 1)
# set Transect to 0
HR_23_hyp$Transect <- 0
# set resp to 1
HR_23_hyp$resp <- 1


# predict conditional fitness estimates for this hypothetical individual
HR_23_hyp.p <- predict(HR23_g2, varvar = varb, idvar = id, root = root,
  newdata = HR_23_hyp, model.type = "conditional")

# take the product of these conditional estimates, EXCEPT for the sub sampling nodes (2nd and 4th node)
mu.fit<-prod(HR_23_hyp.p[c(-2,-4)])
mu.fit
#30.83049
```



## Estimate Mean Fitness and SE
```{r}
#use poisson nodes 

#conditional
nnode <- length(vars)
nind <- length(unique(HR_23_hyp$id))
nnode * nind == nrow(HR_23_hyp)

pout.cond <- predict(HR23_g2, newdata = HR_23_hyp, varvar = varb, idvar = id, root = root, model.type = "conditional", is.always.parameter = TRUE, gradient= TRUE)
xi<-pout.cond$fit

class(xi)
length(xi)==nind*nnode

xi <- matrix(xi, nrow = nind)
colnames(xi) <- vars
xi

#unconditional
pout.unco<-predict(HR23_g2, newdata = HR_23_hyp, varvar = varb, idvar = id, root = root, gradient=TRUE)
mu<-pout.unco$fit
mu <- matrix(mu, nrow = nind)
colnames(mu) <- vars
mu


is.seeds<-grep("filled_seeds", vars)
is.fruits<-grep("total_fruits", vars)

is.seeds
is.fruits

#unconditional estimate of total fruits 
mu.fruits<-mu[, is.fruits]

#conditional estimate of total seeds 
xi.seeds <- xi[ , is.seeds]

mu.seeds <- mu.fruits * xi.seeds
mu.seeds
# Total_fruits
# 10.16336

#make a  function
foo <- function(x) {
# x is xi and mu strung out as one vector
xi <- x[1:length(xi)]
mu <- x[- (1:length(xi))]
xi <- matrix(xi, nrow = nind)
mu <- matrix(mu, nrow = nind)
mu.fruits <- mu[ , is.fruits]
xi.seeds <- xi[ , is.seeds]
mu.seeds <- mu.fruits * xi.seeds
}

#vector of conditional and unconditional estimates for each node 
ximu <- c(xi, mu)
all.equal(foo(ximu), mu.seeds)

foo(ximu)
mu.seeds
# total_fruits
#   10.16336

library(numDeriv)
# R pHRkage numDeriv figures out the jacobian matrix for this transformation.
#We also need the jacobian matrix for the transformation from the “coefficients” vector to the vector ximu

jac.foo <- jacobian(foo, ximu)
jac.ximu <- rbind(pout.cond$gradient, pout.unco$gradient)
jac.total <- jac.foo %*% jac.ximu
#Now the delta method says the variance-covariance matrix of all the fitnesses (the vector estimate mu.fit) is JI−1JT, where J is the overall jacobian matrix jar.total and I is Fisher information for the “coefficients” vector
#there are six coefficients from the model
V <- jac.total %*% solve(HR23_g2$fisher) %*% t(jac.total)
#and the standard errors are square roots of the variances (the diagonal elements of V)
se <- sqrt(diag(V))
bar.pois <- cbind(as.numeric(mu.seeds), se)
colnames(bar.pois) <- c("Estimate", "SE")
bar.pois
# Estimate       SE
#10.16336 1.199265
```

#GxE Effects

##merge the datasets for GxE analysis
```{r}
library(dplyr)

HR_22_gxe <- HR_22 %>%
  mutate(Year = "2022") %>%
  dplyr::select(Transect, Donor, Recipient, sample_ID, surv_to_flower, F_plant,
         total_fruits, closed_fruits, filled_seeds, Year)

HR_23_gxe <- HR_23 %>%
  mutate(Year = "2023") %>%
  dplyr::select(Transect, Donor, Recipient, sample_ID_SEG, surv_to_flower, F_plant,
         total_fruits, closed_fruits, filled_seeds, Year) %>%
  rename(sample_ID = sample_ID_SEG)

HR_G1_both <- bind_rows(HR_22_gxe, HR_23_gxe) %>%
  mutate(
    Year = factor(Year),
    Donor = factor(Donor),
    Recipient = factor(Recipient)
  )
```

##set up aster
```{r}
# surv_to_flower and F_plant are all-1s among fitness plants at HR 2022 (100%
# survival), causing perfect separation that prevents reaster from converging.
# The GxE question concerns fitness expression, so we use only the 3 terminal
# fitness nodes: total_fruits -> closed_fruits -> filled_seeds.
vars <- c("total_fruits", "closed_fruits", "filled_seeds")

HR_G1_both <- as.data.frame(HR_G1_both)

HR_G1_aster <- reshape(
  HR_G1_both, varying = list(vars), direction = "long", timevar = "varb", times = as.factor(vars), v.names = "resp")

#####Check that the reshape worked
#If it did the number of rows will be the same
nrow(HR_G1_aster)
nrow(HR_G1_both)*length(vars)

fit <- grepl("filled_seeds", as.character(HR_G1_aster$varb))
fit <- as.numeric(fit)
HR_G1_aster$fit <- fit

#set up root that is year specific
HR_G1_aster$root <- ifelse(HR_G1_aster$Year == "2022", 8, 5) #This is important bc the root changes in 2023 vs 2022!

pred <- c(0,1,2)
fam  <- c(2,1,2)

#check that the distribution you assigned is correct
sapply(fam.default(), as.character)[fam]

#show graphical model
foo <- c("root", vars)
pvars <- foo[pred +1]
bar <-cbind(pvars, vars)
colnames(bar) <- c("pred", "succ")
```

##set up parental and random effects
```{r}
#parental effect
modmat.Donor <- model.matrix(~ 0 + fit:Donor, HR_G1_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, HR_G1_aster)
modmat.donorrecipient <- cbind(modmat.Donor, modmat.Recipient)

#GxE random effect
modmat.parental.year <- model.matrix(~ 0 + fit:Donor:Year, HR_G1_aster)
modmat.recipient.year <- model.matrix(~ 0 + fit:Recipient:Year, HR_G1_aster)
modmat.parentalyear <- cbind(modmat.parental.year, modmat.recipient.year)
```

##run both models
```{r}
#fit model WITHOUT GxE
rout_no_gxe <- reaster(
  resp ~ fit + varb + fit:(Year + Transect),
  random = list(Parental = ~ 0 + modmat.donorrecipient),
  pred = pred,
  fam = fam,
  varvar = varb,
  idvar = id,
  root = root,
  data = HR_G1_aster
)

#fit model WITH GxE

rout_gxe <- reaster(
  resp ~ fit + varb + fit:(Year + Transect),
  random = list(
    Parental = ~ 0 + modmat.donorrecipient,
    ParentalYear = ~ 0 + modmat.parentalyear
  ),
  pred = pred,
  fam = fam,
  varvar = varb,
  idvar = id,
  root = root,
  data = HR_G1_aster
)
```


##model comparison
```{r}
##################################################################
#Likelihood ratio test for effect of Donor X Year interaction
##################################################################
gxy_lrt <- anova(rout_no_gxe, rout_gxe)
gxy_lrt
summary(rout_gxe)

# Boundary-corrected p-value (Self & Liang 1987): H0: sigma^2(ParentalYear) = 0
# lies on the boundary, so the reference distribution is 50:50 chi^2_0 + chi^2_1.
# Halving the standard LRT p-value gives the correct one-sided boundary test.
cat("\nBoundary-corrected G×Y p:", gxy_lrt$`Pr(>Chisq)`[2] / 2, "\n")
```


## Between-year sire correlation with shrinkage diagnostic (#6)
```{r sire-correlation-shrinkage}
# Load both years of BLUPs for this site. The Pearson r is influenced by
# year-specific shrinkage asymmetry (sigma differs between years), which
# can create apparent rank changes even when true breeding values are
# stable. Spearman r is more robust. We also report sigma_Y1/sigma_Y2 as
# a shrinkage-asymmetry diagnostic.

site_code <- sub("Aster6_([A-Za-z]+)\.Rmd", "\1",
                 knitr::current_input())

# resolve to 2-letter code
site_map <- c(Angelo="AC", BlueOak="BO", BodegaBay="BB", Hastings="HR")
site <- site_map[site_code]

bv22_path <- paste0("processed_data/", site, "22_bhat.Donor.mu.csv")
bv23_path <- paste0("processed_data/", site, "23_bhat.donor.mu.csv")

bv22 <- read.csv(bv22_path)
bv23 <- read.csv(bv23_path)
names(bv22)[1:2] <- c("raw", "BV_Y1")
names(bv23)[1:2] <- c("raw", "BV_Y2")
bv22$Donor <- sub(".*Donor", "", bv22$raw)
bv23$Donor <- sub(".*Donor", "", bv23$raw)
bv_both <- merge(bv22[, c("Donor","BV_Y1")],
                 bv23[, c("Donor","BV_Y2")], by = "Donor")

pearson  <- cor.test(bv_both$BV_Y1, bv_both$BV_Y2, method = "pearson")
spearman <- cor.test(bv_both$BV_Y1, bv_both$BV_Y2, method = "spearman",
                     exact = FALSE)

# Sigma values were computed earlier in this script
sig_y1 <- rout_HR22_2$sigma["Parental"]
sig_y2 <- rout_HR23_2$sigma["Parental"]

cat("
--- Between-year sire correlation:", site, "---
")
cat("n donors with both years:", nrow(bv_both), "
")
cat("Pearson  r =", round(pearson$estimate, 3),
    " p =", round(pearson$p.value, 4), "
")
cat("Spearman rho =", round(spearman$estimate, 3),
    " p =", round(spearman$p.value, 4), "
")
cat("sigma_Y1:", round(sig_y1, 5),
    "  sigma_Y2:", round(sig_y2, 5),
    "  ratio Y1/Y2:", round(sig_y1/sig_y2, 2), "
")
cat("(ratio >> 1 or << 1 indicates shrinkage asymmetry that biases Pearson r)
")
```

