---
title: "Aster6_BodegaBay"
author: "Helen Payne"
date: "2025-07-30"
output:
  pdf_document: default
  html_document: default
---

# Bodega Bay Generation 1 2022

Load in data
```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
library(here)

#Read in the data
BB_22 <- read_csv(here::here("data/BB_compiledsheet_full_2022.csv")) 

#organize the data
BB_22 <- BB_22 %>%
  mutate(
    sample_ID = str_c(Year, Recipient, Gen, str_pad(Location, width = 3, pad = "0"), Plant_ID, sep = "-"), #create an id column for each individual
    sample_ID = str_replace(sample_ID, "^([^_]*)_", "\\1-"), #use "_" rather than "-"
    F_plant = ifelse(F_plant == TRUE, 1, 0), #make fitness plant binary (1/0) rather than true/false
    Location = as.numeric(Location)) #make sure location is a continuous variable

#clean the data
BB_22 <- BB_22 %>%
  mutate(
    surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower), #If a plant produced any fruits, it must have survived to flowering, even if the original surv_to_flower entry was coded as 0, This corrects miscodings where survival to flowering was mistakenly marked as 0 despite having fruit production
  ) %>%
  dplyr::select("Location", "Donor", "Recipient", "sample_ID", "surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds") %>% #select the columns we want
  filter(!is.na(Location)) # We are removing W1 and W2 from these analyses because they were part of a different experiment involving a watered treatment. 
## In the original dataset, we included some extra fitness plants (see `anyFit_P`) in other transects for comparison with W1 and W2. 
### For the purposes of these analyses, we will exclude them and focus only on the `F_plant` values that are recorded for every segment.

BB_22$F_plant[is.na(BB_22$F_plant)] <- 0 #we need to make F_plant 0 and not NA

#Now we want to make sure every segment is represented by one F_plant (because we will filter by F_plant soon)
BB_22 <- BB_22 %>%
  mutate(sample_prefix = substr(sample_ID, 1, 19)) %>% #create a sample prefix with the first 19 characters of sample ID
  group_by(sample_prefix) %>% #group by the prefix
  mutate(
    all_dead = all(surv_to_flower == 0), #flags groups where no plants survived and where there are currently no fitness plants
##if a group has no fitness plants, it randomly assigns one plant in the group to be the fitness plant
    no_f_plant = sum(F_plant) == 0,
    F_plant = if_else(
      no_f_plant & row_number() == sample(n(), 1),
      1L,
      F_plant
    ) #this will pick different individuals within the group each time unless you set a seed. For the purposes of this analyses, it will not matter because their values will be zero either way.
  ) %>%
  ungroup() %>%
  dplyr::select(-sample_prefix, -all_dead, -no_f_plant) #ungroup and remove the rows that were created

#Now we want to determine how many individuals within a segment survived to flower
BB_22 <- BB_22 %>%
  mutate(group = ceiling(row_number() / 8)) %>%  # Create a grouping variable for every 8 rows
  # Calculate the sum of 'surv_to_flower' within each group
  group_by(group) %>%
  mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>% # Ungroup to avoid grouped data frame
  # Remove the temporary 'group' column
  dplyr::select(-group)


#Filter to just fitness plants
BB_22_filtered <- BB_22 %>%
  filter(F_plant == 1)

#Reorder for compatibility
BB_22 <- BB_22_filtered %>%
  dplyr::select(Location, Donor, Recipient, sample_ID, total_surv_to_flower, F_plant, total_fruits, closed_fruits, filled_seeds)

#Rename for compatibility
BB_22 <- BB_22 %>%
  rename(surv_to_flower = total_surv_to_flower)

#change F_plant to 0 if survival is 0
##this will allow aster to function- F_plant cannot be 1 if survival is 0. 
BB_22 <- BB_22 %>%
  mutate(F_plant = ifelse(surv_to_flower == 0, 0, F_plant))
```

```{r}
#check for nonsense data
subset(BB_22, filled_seeds > 0 & surv_to_flower == 0) #If a plant did not survive to flower, are there still filled seeds recorded? no.
subset(BB_22, filled_seeds > 0 & total_fruits == 0) #If a plant did not have fruit, are there still filled seeds recorded? no.
subset(BB_22, filled_seeds > 0 & closed_fruits == 0) #If a plant did not have closed fruits, are there still filled seeds recorded? no.

#Make Donors and Recipients factors
BB_22$Donor <- as.factor(BB_22$Donor)
BB_22$Recipient <- as.factor(BB_22$Recipient)

#number that germinated from table 1
BB_22 %>%
  mutate(surv_to_flower = as.factor(surv_to_flower)) %>% 
  group_by(surv_to_flower) %>% 
  count()

#distinct number of Recipients
BB_22 %>%
  distinct(Recipient) %>%
  count()
#147

#distinct number of Donors
BB_22  %>%
  distinct(Donor) %>%
  count()
#49

#number of Donors, and number of individuals for each recipient
NDS<-BB_22 %>%
  group_by(Recipient, Donor) %>% 
  count()

NDS #they should have between 1-3
```

```{r}
#make NAs for life history stages zero
BB_22$surv_to_flower[is.na(BB_22$surv_to_flower)] <- 0
BB_22$total_fruits[is.na(BB_22$total_fruits)]<-0
BB_22$closed_fruits[is.na(BB_22$closed_fruits)]<-0
BB_22$filled_seeds[is.na(BB_22$filled_seeds)]<-0

# Define the update_zeros function modifies the data and fixes cases where there are any inconsistencies; because the data looks good, it shouldn't change anything for this df
update_zeros <- function(df) {
  for (i in 1:nrow(df)) {
    if (df$surv_to_flower[i] == 0) {
      df$total_fruits[i] <- 0
      df$closed_fruits[i] <- 0
      df$filled_seeds[i] <- 0
    }
    if (df$total_fruits[i] == 0) {
      df$closed_fruits[i] <- 0
      df$filled_seeds[i] <- 0
    }
    if (df$closed_fruits[i] == 0) {
      df$filled_seeds[i] <- 0
    }
  }
  return(df)
  }

# Apply the function to the dataset
BB_22 <- update_zeros(BB_22)

#save it
write.csv(BB_22, "tables/BB_22.csv", row.names = FALSE)
```

```{r}
#Check the distribution of zeros

# Number of segments with zero survival
surv_to_flower_zero <- sum(BB_22$surv_to_flower == 0)
cat("Number of rows where surv_to_flower is 0:", surv_to_flower_zero, "\n")
#78

# Determine how many rows are equal to 0 for the column "total_fruits"
total_fruits_zero <- sum(BB_22$total_fruits == 0)
cat("Number of rows where total_fruits is 0:", total_fruits_zero, "\n")
#283

# Determine how many rows are equal to 0 for the column "closed_fruits"
closed_fruits_zero <- sum(BB_22$closed_fruits == 0)
cat("Number of rows where closed_fruits is 0:", closed_fruits_zero, "\n")
#308

# Determine how many rows are equal to 0 for the column "filled_seeds"
filled_seeds_zero <- sum(BB_22$filled_seeds == 0)
cat("Number of rows where filled_seeds is 0:", filled_seeds_zero, "\n")
#346
```

## Set up Aster
```{r}
library(aster)
library(reshape2)

#the "vars" are the nodes in the aster graphical model.   
vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

#convert to a dataframe for reshape to work
BB_22 <- as.data.frame(BB_22)

#Reshape BB to longform. "varying" notes the set of variables that will be converted to longform and named in "timevar" under one column of varb. "Times" says to use vars for the newly created column varb 
BB22_aster <- reshape(BB_22 ,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#####Check that the reshape worked 
#If it did the number of rows will be the same 
nrow(BB22_aster)
nrow(BB_22)*length(vars)

#####Designate fitness variable, "filled_seeds", which is the final node 
#grep1 is like a search function
fit<-grepl("filled_seeds", as.character(BB22_aster$varb))
fit<-as.numeric(fit)
BB22_aster$fit <- fit

#####Check "filled_seeds", is designated as the fitness variable 
with(BB22_aster, sort(unique(as.character(varb)[fit==0])))
with(BB22_aster, sort(unique(as.character(varb)[fit==1])))

#####Add "root" to BB22_aster files where value is 1 
BB22_aster <- data.frame(BB22_aster, root=8)

#####Set graphical node and distribution for fitness nodes(preds) 
names(BB22_aster)

#pred is nodes. There are 5 nodes, so five preds.
pred <- c(0,1,2,3,4)

#fam assigns a distribution to each node. 1= bernoulli, 2= poission. The subsampling nodes (node #2 and #4 in pred) are binomial so Bernouli (always). Bernouli is a specific type of binomial so the assignment for family distribution is the same
fam<-c(1,1,2,1,2)
```



## Random Effects Aster Models
```{r}
#Run a model with both donor and recipient
Both_BB22 <- reaster(resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
    fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = BB22_aster)
summary(Both_BB22)

# Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
#     fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, 
#     fam = fam, varvar = varb, idvar = id, root = root, data = BB22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.2716488  0.1419454  -8.959  < 2e-16 ***
# fit                 2.4775688  0.1659669  14.928  < 2e-16 ***
# varbF_plant         0.1991739  0.1605987   1.240    0.215    
# varbsurv_to_flower  1.1817996  0.1465994   8.061 7.54e-16 ***
# varbtotal_fruits    0.8437106  0.2070647   4.075 4.61e-05 ***
# fit:Location       -0.0013273  0.0002672  -4.967 6.79e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#           Estimate Std. Error z value Pr(>|z|)/2    
# Donor      0.00000         NA      NA         NA    
# Recipient  0.12905    0.01513   8.528     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


#Check for the effect of Donor alone. Significant. 
Donor_BB22<- reaster(resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
    fit:Donor), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = BB22_aster)
summary(Donor_BB22)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
#     fit:Donor), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = BB22_aster)
# 
# 
# Fixed Effects:
#                     Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.310945   0.142211  -9.218  < 2e-16 ***
# fit                 2.550519   0.165757  15.387  < 2e-16 ***
# varbF_plant         0.215194   0.161422   1.333    0.182    
# varbsurv_to_flower  1.221095   0.146857   8.315  < 2e-16 ***
# varbtotal_fruits    0.883007   0.207247   4.261 2.04e-05 ***
# fit:Location       -0.001374   0.000265  -5.185 2.16e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#       Estimate Std. Error z value Pr(>|z|)/2    
# Donor  0.09704    0.01622   5.984   1.09e-09 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


####Testing for Donor and Recipient significance
Recipient_BB22 <- reaster(resp ~ fit +  varb + fit:Location, random = list(Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = BB22_aster)
summary(Recipient_BB22)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Recipient = ~0 + 
#     fit:Recipient), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = BB22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.2716464  0.1419454  -8.959  < 2e-16 ***
# fit                 2.4775633  0.1659670  14.928  < 2e-16 ***
# varbF_plant         0.1991733  0.1605987   1.240    0.215    
# varbsurv_to_flower  1.1817973  0.1465994   8.061 7.54e-16 ***
# varbtotal_fruits    0.8437085  0.2070647   4.075 4.61e-05 ***
# fit:Location       -0.0013273  0.0002672  -4.967 6.79e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#           Estimate Std. Error z value Pr(>|z|)/2    
# Recipient  0.12905    0.01513   8.528     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


####Combine donor and recipient into one random effect, `Parental`. The estimates are of similar magnitude 
modmat.Donor <- model.matrix(~ 0 + fit:Donor, BB22_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, BB22_aster)
modmat.donorrecipient <- cbind(modmat.Donor,modmat.Recipient)

rout_BB22 <- reaster(resp ~ fit + varb, list(Parental = ~ 0 + modmat.donorrecipient),pred, fam, varb, id, root, data = BB22_aster)
sout_BB22 <-summary(rout_BB22)
sout_BB22

#Call:
# reaster.formula(fixed = resp ~ fit + varb, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = BB22_aster)
# 
# 
# Fixed Effects:
#                    Estimate Std. Error z value Pr(>|z|)    
# (Intercept)         -1.2956     0.1422  -9.113  < 2e-16 ***
# fit                  2.4167     0.1669  14.482  < 2e-16 ***
# varbF_plant          0.2091     0.1611   1.297    0.194    
# varbsurv_to_flower   1.2057     0.1468   8.213  < 2e-16 ***
# varbtotal_fruits     0.8676     0.2072   4.187 2.83e-05 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental  0.09590    0.01208   7.938   1.03e-15 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#model comparison for significance of Location. Significant.
rout_BB22_2 <- reaster(resp ~ fit + varb + fit:Location, list(Parental = ~ 0 + modmat.donorrecipient),
    pred, fam, varb, id, root, data = BB22_aster)
sout_BB22_2 <-summary(rout_BB22_2)
sout_BB22_2

save(rout_BB22_2, file = "processed_data/rout_BB22_2.rdata")

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = BB22_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.2755413  0.1419863  -8.984  < 2e-16 ***
# fit                 2.4857110  0.1663310  14.944  < 2e-16 ***
# varbF_plant         0.2007041  0.1606814   1.249    0.212    
# varbsurv_to_flower  1.1856922  0.1466390   8.086 6.18e-16 ***
# varbtotal_fruits    0.8476033  0.2070927   4.093 4.26e-05 ***
# fit:Location       -0.0013497  0.0002688  -5.022 5.12e-07 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental  0.09282    0.01189   7.806   2.95e-15 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova(rout_BB22, rout_BB22_2)

#Analysis of Deviance Table

# Model 1: resp ~ fit + varb, ~0 + modmat.donorrecipient
# Model 2: resp ~ fit + varb + fit:Location, ~0 + modmat.donorrecipient
#   Mod Df Fix Mod Df Rand Mod Dev Df Fix Df Rand Deviance    P-value
# 1          5           1 -7712.6                                   
# 2          6           1 -7686.2      1       0   26.452 2.7011e-07

####Now look at parameter estimates to be sure we have the model intended.
sout_BB22_2 <-summary(rout_BB22_2)
sout_BB22_2
```


## Mapping function
```{r}
#following Geyer et al. 2022
library(numDeriv)
fishinv<-solve(sout_BB22_2$fisher)
map.factory <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  
  #extracts the embeded aster object from the reaster result without the random effects- aout is the fixed effect aster model
  aout <- rout$obj
  
  #confirms it is an aster object
  stopifnot(inherits(aout, "aster"))
  
  #the number of nodes in the aster graph
  nnode <- ncol(aout$x)
  
  # makes sure is.subsamp matches the number of nodes
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  
  #vector of fixed effect coefficients from the fitted model on the canonical scale
  alpha <- rout$alpha
  
  #error if "fit" isn't found or isn't unique
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  
  #get the model matrix (a design matrix) for the first individual only. It is a three dimensional array that encodes the relationship between: individuals x nodes x fixed-effect predictors. modmat[i, j, k] = the value of the k-th fixed effect predictor for individual i at node j. For example, modmat[1, 2, 3] might be the value of predictor 3 (say, "Location") for node 2 (say, flowering) of individual 1. We can use just one individual's design matrix to compute predictions if we’re assuming all individuals have the same covariate values — or if we override the covariates (like setting location = 0) and are only interested in mean fitness, not individual-level variation.
  modmat <- rout$obj$modmat[1, , , drop = FALSE] 
  
  #set the effect of "Location" to zero on the canonical scale, pretend all individuals have the same location
  modmat[ , , "fit:Location"] <- 0
  
  # This creates the root array in the form expected by the aster model. set root = 8, dim = dim(modmat)[1:2] selects the first two dimensions of the 3Dmodmat array ,which is the shape of the root argument expected by aster
  root <- array(8, dim = dim(modmat)[1:2]) 
  
  #Extract the response vector for the first individual, across all nodes, there are five nodes. aout is the fixed effect aster model, with no random effects, and x are your observed values, not predicted values. This line of code extracts your observed values for the first individual for each of the five nodes. But you’re not using it for its actual numbers — you're using it as a template to define the node types and graph structure which is used for the prediction
  x <- aout$x[1, c(1:5)]
  print(x) #check that the X in the mapping function is indeed giving you your design matrix with just the nodes and nothing else
  
  # # return map function
  #This part returns a function that accepts a numeric input b (a shift on the canonical scale), adjusts the model accordingly, and outputs the predicted mean fitness:
  function (b) {
    #error checks to ensure "b" is a valid numeric scalar- ask what is b
    stopifnot(is.numeric(b))
    stopifnot(is.finite(b))
    stopifnot(length(b) == 1)
    
    #Adjust the "fit" coefficient by b. This moves you along the canonical scale.
    alpha[ifit] <- alpha[ifit] + b
    
    #Predicts conditional expectations (xi) for each node under the new coefficient vector.
    xi <- predict(
      aout,
      x = x, 
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      
      #ensures that all nodes are treated as parameters.
      is.always.parameter = TRUE)
  
   # Reshape the predicted values to a matrix.
    #Drop subsampled nodes (if any), so we're focusing on unconditional predictions.
    xi <- matrix(xi, ncol = nnode)
    # always use drop = FALSE unless you are sure you don't want that
    # here if we omit drop = FALSE and there is only one non-subsampling
    # node, the code will break (apply will give an error)
    xi <- xi[,!is.subsamp, drop = FALSE]
    
    #multiply conditional expectations along the graph to get the unconditional expectation at the end node. This takes the product of the conditional expectations along the path for each individual. In aster theory, that product gives you the unconditional mean of the terminal node (i.e., mean fitness), under the model.
    mu <- apply(xi, 1, prod) #every individual
    # mu is unconditional mean values for model without subsampling
    # in this application all components mu are the same because no
    # covariates except varb, so just return only one
    mu[1] 
  }
}

map <- map.factory(rout_BB22_2, vars %in% c("F_plant", "closed_fruits"))

#MEAN FITNESS
#centers on zero on the canonical scale 
mf<-map(0)
mf
#1.486183

#ADDITIVE GENETIC VARIANCE FOR FITNESS
vaw <- rout_BB22_2$sigma["Parental"]^2 * grad(map, 0)^2
# get rid of name
vaw <- as.numeric(vaw)
vaw <- vaw * 4
vaw
#1.676352

####FFTNS
fftns <- vaw / mf
fftns
#1.127958

#prediction of progeny generation mean fitness
mf+fftns
#2.614141

# get SE
map.factory.too <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  # # modmat for one individual. Taking the first row of modmat- there is one row for each individual
   modmat <- rout$obj$modmat[1, , , drop = FALSE] #this only takes the first row 
   modmat[ , , "fit:Location"] <- 0
   # # set root = 8
   root <- array(8, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, second number is number of nodes 
  # return map function
  function (balpha) {
    stopifnot(is.numeric(balpha))
    stopifnot(is.finite(balpha))
    stopifnot(length(balpha) == 1 + length(alpha))
    b <- balpha[1]
    alpha <- balpha[-1]
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
       x = x, # added
      newcoef = alpha,
       modmat = modmat,
       root = root,
      model.type = "conditional",
      is.always.parameter = TRUE
    )
    xi <- matrix(xi, ncol = nnode)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod)
    # mu is unconditional mean values for model without subsampling
    mu[1]
  }
}

map.too <- map.factory.too(rout_BB22_2, vars %in% c("F_plant", "closed_fruits"))


#centering donor on zero
balpha.hat <- c(0, rout_BB22_2$alpha)

map.too(balpha.hat)
#1.486183

all.equal(map(0), map.too(balpha.hat))

#first and second derivatives
g<-grad(map.too, balpha.hat)

#calculates the nxn matrix
h <- hessian(map.too, balpha.hat)

#these are partial derivatives
dmu.db <- g[1]
dmu.dalpha <- g[-1]
d2mu.db.dalpha <- h[1, -1]

# give names to the estimators in our formulas
mu.hat <- map.too(balpha.hat)
nu.hat <- rout_BB22_2$nu["Parental"]

#FFTNS SE
#calculations for the gradient vector of the FFTNS prediction with respect to the parameters of the models (fixed effects and variance components) 
#zero removed
dfftns <- c(- 4 * nu.hat * dmu.dalpha * dmu.db^2 / mu.hat^2 +
8 * nu.hat * d2mu.db.dalpha / mu.hat, 4 * dmu.db^2 / mu.hat)

#And apply the delta method.
fftns.se <- t(dfftns) %*% fishinv %*% dfftns
fftns.se <- sqrt(as.vector(fftns.se))
fftns.se
#0.2926037

#ADDITIVE GENETIC VARIANCE SE
#calculate the gradient vector VA(W) with respect to the parameters of the model using these formulae
dvaw <- c(8 * nu.hat * d2mu.db.dalpha, 4 * dmu.db^2)

#And then apply the delta method to get standard errors for this estimator
vaw.se <- t(dvaw) %*% fishinv %*% dvaw
vaw.se <- sqrt(as.vector(vaw.se))
vaw.se
#0.3710638

#MEAN FITNESS SE
dmf <- c(dmu.dalpha, 0)
mf.se <- t(dmf) %*% fishinv %*% dmf
mf.se <- sqrt(as.vector(mf.se))
mf.se
#0.1853252
```

## Breeding Values

```{r}
#extract breeding values
bhat <- rout_BB22_2$b 
bhat.Donor<- bhat[grep("Donor", names(bhat))]

#check this is the correct number of donors (49)
length(bhat.Donor)

#convert canonical values to mean value parameter values using mapping function.
map <- map.factory(rout_BB22_2, vars %in% c("F_plant", "closed_fruits"))
vectorized.map <- Vectorize(map)
curve(vectorized.map, from = -1 / 2, to = 1/2, xlab="b", ylab=expression(mu(b)))

#breeding values for donor groups on the mean value parameter scale 
bhat.Donor.mu <- vectorized.map(bhat.Donor)
bhat.Donor.mu<-as.data.frame(bhat.Donor.mu)
dmap_BB22 <- Vectorize(function(b) numDeriv::grad(map, b))
bhat.Donor.mu$BV_se_approx <- abs(dmap_BB22(bhat.Donor)) *
  rout_BB22_2$sigma["Parental"]
write.csv(bhat.Donor.mu, "tables/BB22_bhat.Donor.mu.csv")

#plot the distribution of breeding values
# Extract the numeric column from bhat.Donor.mu
numeric_values <- bhat.Donor.mu[["bhat.Donor.mu"]]  # Column name is the same as the data frame name

# Calculate and plot the density
prob_den <- density(numeric_values)
plot(prob_den)
```

```{r}
#BACK OF NAPKIN FITNESS CALCULATION BB 2022 G1

W <- BB_22 %>%
  summarise(
    total_seeds = sum(filled_seeds, na.rm = TRUE),
    W = total_seeds / n()
  )

W
#2.300207
```

# Bodega Bay Generation 1 2023
```{r}
#load year 2 (2023) life history data for BB site
BB_23 <- read_csv(here::here("data/BB_compiledsheet_full_2023.csv")) 

#organize the data
BB_23 <- BB_23 %>%
  dplyr::select("Location", "Gen", "Donor", "Recipient", "sample_ID_SEG", "surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds", "msm_all") %>%
  mutate(F_plant = ifelse(F_plant == TRUE, 1, 0)) %>%
  mutate(surv_to_flower = ifelse(total_fruits > 0 & surv_to_flower == 0, 1, surv_to_flower)) #if a plant made fruits, it must have survived to flowering, so the survival indicator is changed from 0 to 1

BB_23 <- BB_23 %>%
  # Create a grouping variable
  mutate(group = sample_ID_SEG) %>%
  # Calculate the sum of 'surv_to_flower' within each group
  group_by(group) %>%
  mutate(total_surv_to_flower = sum(surv_to_flower, na.rm = TRUE)) %>%
  ungroup() %>% # Ungroup to avoid grouped data frame
  # Remove the temporary 'group' column using dplyr::select
  dplyr::select(-group)

#Ensure at least one F_plant per sample_ID prefix (first 19 characters)
BB_23 <- BB_23 %>%
  group_by(sample_ID_SEG) %>%
  arrange(Location, Recipient, Donor, sample_ID_SEG, .by_group = TRUE) %>%
  mutate(
    has_f = any(F_plant == 1),
    m_surv = sum(surv_to_flower == 1, na.rm = TRUE),
    # if none marked, pick deterministically:
    F_plant = case_when(
      has_f ~ F_plant,# keep existing
      !has_f & m_surv == 0 & row_number() == 1L ~ 1L,    # no survivors: pick first row (placeholder)
      TRUE ~ F_plant
    )
  ) %>%
  ungroup() %>%
  dplyr::select(-has_f, -m_surv)


#Filter to just fitness plants
BB_23 <- BB_23 %>%
  filter(F_plant == 1)

BB_23_rearranged <- BB_23 %>%
  dplyr::select(Location, Gen, Donor, Recipient, sample_ID_SEG, total_surv_to_flower, F_plant, total_fruits, closed_fruits, filled_seeds, msm_all)


BB_23_rearranged <- BB_23_rearranged %>%
  rename(surv_to_flower = total_surv_to_flower)


#change F_plant to 0 if survival is 0 to let aster work
BB_23 <- BB_23_rearranged %>%
  mutate(F_plant = ifelse(surv_to_flower == 0, 0, F_plant))
```


```{r}
####make NAs for other life history stages zero
BB_23[is.na(BB_23$surv_to_flower),]$surv_to_flower <- 0
BB_23[is.na(BB_23$total_fruits),]$total_fruits <- 0
BB_23[is.na(BB_23$closed_fruits),]$closed_fruits <- 0
BB_23[is.na(BB_23$filled_seeds),]$filled_seeds <- 0


####check for nonsense data
subset(BB_23, filled_seeds > 0 & surv_to_flower  == 0)
subset(BB_23, filled_seeds > 0 & total_fruits == 0)
subset(BB_23, filled_seeds > 0 & closed_fruits ==0)
subset(BB_23, filled_seeds > 0 & filled_seeds ==0)
subset(BB_23, closed_fruits > total_fruits)
subset(BB_23, total_fruits > 0 & surv_to_flower  == 0)


#make factors
BB_23$Donor <- as.factor(BB_23$Donor)
BB_23$Recipient <- as.factor(BB_23$Recipient)
BB_23$Gen<-as.factor(BB_23$Gen)

```

## Germination in G1 vs G2 (Parental vs Progeny)

```{r}
#data for table 1
#counts the number of Recipients in parental and progeny generation
BB_23 %>%
  group_by(Gen) %>% 
  distinct(Recipient) %>%
  arrange(Gen, Recipient) %>%
  count()

#counts the number of Donors in each group. For progeny number of donors is grand-Donors as they were open pollinated in the field 
BB_23 %>%
  group_by(Gen) %>% 
  distinct(Donor) %>%
  count()

BB_23 %>%
  group_by(Gen, Recipient, Donor) %>% 
  count()
```

```{r}
#Check the distribution of zeros in your model

# Determine how many rows are equal to 0 for the column "surv_to_flower"
surv_to_flower_zero <- sum(BB_23$surv_to_flower == 0)
cat("Number of rows where surv_to_flower is 0:", surv_to_flower_zero, "\n")
#136

# Determine how many rows are equal to 0 for the column "total_fruits"
total_fruits_zero <- sum(BB_23$total_fruits == 0)
cat("Number of rows where total_fruits is 0:", total_fruits_zero, "\n")
#243

# Determine how many rows are equal to 0 for the column "closed_fruits"
closed_fruits_zero <- sum(BB_23$closed_fruits == 0)
cat("Number of rows where closed_fruits is 0:", closed_fruits_zero, "\n")
#258

# Determine how many rows are equal to 0 for the column "filled_seeds"
filled_seeds_zero <- sum(BB_23$filled_seeds == 0)
cat("Number of rows where filled_seeds is 0:", filled_seeds_zero, "\n")
#291

```

## Seed weight
```{r}
#regress fitness on seed weight
seedWgt_BB <- BB_23 %>%
  dplyr::select(sample_ID_SEG, Recipient, Gen, surv_to_flower, msm_all, filled_seeds) %>%
  filter(!is.na(msm_all)) %>%
  group_by(Recipient) 

# t-test of seed weight between generations
t.test(subset(seedWgt_BB, Gen == "G1")$msm_all, subset(seedWgt_BB, Gen == "G2")$msm_all)

#	Welch Two Sample t-test
# 
# data:  subset(seedWgt_BB, Gen == "G1")$msm_all and subset(seedWgt_BB, Gen == "G2")$msm_all
# t = -0.90345, df = 191.12, p-value = 0.3674
# alternative hypothesis: true difference in means is not equal to 0
# 95 percent confidence interval:
#  -0.6155446  0.2288039
# sample estimates:
# mean of x mean of y 
#  3.615985  3.809355 
 
#average seed weight of progeny and parental
seedWgt_BB%>%
  group_by(Gen) %>%
  summarise(Average=mean(msm_all))
```


## Set up aster
```{r}
#subset the data to only Gen 1
BB23_aster_a<-subset(BB_23, Gen=="G1")

BB23_aster_a <- BB23_aster_a[, !names(BB23_aster_a) %in% "msm_all"]

#set the nodes for the aster graphical model as "vars".
vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

BB23_aster_a <- as.data.frame(BB23_aster_a)

#reshape BB23_aster_a to longform
BB23_aster <- reshape(BB23_aster_a,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#Check that the reshape worked 
nrow(BB23_aster)
nrow(BB23_aster_a)*length(vars)

#designate fitness variable as terminal node, "filled_seeds"
fit<-grepl("filled_seeds", as.character(BB23_aster$varb))
fit<-as.numeric(fit)
BB23_aster$fit<-fit

#check that "filled_seeds" is designated as the fitness variable
with(BB23_aster, sort(unique(as.character(varb)[fit==0])))
with(BB23_aster, sort(unique(as.character(varb)[fit==1])))

#Add a variable "root" to BB23_aster where value is 1
BB23_aster <- data.frame(BB23_aster, root=5)

#give each node a number
pred <- c(0,1,2,3,4)

#designate each node with a distribution. 1= Bernoulli, 2= poisson. sub sampling nodes (2 and 4) is binomial(Bernoulli is are a type of binomial distribution so the family assignment is the same)
fam<-c(1,1,2,1,2)
```

## Random effects aster model
```{r}
####Fit random effects model. donor estimate is zero with recipient  
Both_BB23<- reaster(resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
    fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = BB23_aster)
summary(Both_BB23)


#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
#     fit:Donor, Recipient = ~0 + fit:Recipient), pred = pred, 
#     fam = fam, varvar = varb, idvar = id, root = root, data = BB23_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.3577747  0.0723778 -18.760   <2e-16 ***
# fit                 2.7727427  0.0800207  34.650   <2e-16 ***
# varbF_plant        -4.2738165  0.1701548 -25.117   <2e-16 ***
# varbsurv_to_flower  1.0888217  0.0890951  12.221   <2e-16 ***
# varbtotal_fruits    2.1006560  0.1154331  18.198   <2e-16 ***
# fit:Location       -0.0009951  0.0001063  -9.358   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#           Estimate Std. Error z value Pr(>|z|)/2    
# Donor     0.013601   0.008442   1.611     0.0536 .  
# Recipient 0.035131   0.004659   7.541   2.34e-14 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#Check for the effect of donor alone. Significant. 
donor_BB23<- reaster(resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
    fit:Donor), pred = pred, fam = fam, varvar = varb, 
    idvar = id, root = root, data = BB23_aster)
summary(donor_BB23)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Donor = ~0 + 
#     fit:Donor), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = BB23_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.3632876  0.0724034 -18.829   <2e-16 ***
# fit                 2.7850893  0.0799901  34.818   <2e-16 ***
# varbF_plant        -4.3928149  0.1698471 -25.863   <2e-16 ***
# varbsurv_to_flower  1.0943345  0.0891159  12.280   <2e-16 ***
# varbtotal_fruits    2.1061688  0.1154492  18.243   <2e-16 ***
# fit:Location       -0.0009142  0.0000965  -9.474   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#       Estimate Std. Error z value Pr(>|z|)/2    
# Donor 0.024923   0.003807   6.547   2.94e-11 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#check for the effect of recipient alone. significant  
recipient_BB23<- reaster(resp ~ fit + varb + fit:Location, random = list(Recipient = ~0 + 
    fit:Recipient), pred = pred, fam = fam, varvar = varb, #in model recipient_ML19 shouldn't this say "fit:recipient"
    idvar = id, root = root, data = BB23_aster)
summary(recipient_BB23)

#Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Recipient = ~0 + 
#     fit:Recipient), pred = pred, fam = fam, varvar = varb, idvar = id, 
#     root = root, data = BB23_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.3577981  0.0723779 -18.760   <2e-16 ***
# fit                 2.7726757  0.0800052  34.656   <2e-16 ***
# varbF_plant        -4.2744619  0.1701678 -25.119   <2e-16 ***
# varbsurv_to_flower  1.0888449  0.0890952  12.221   <2e-16 ***
# varbtotal_fruits    2.1006792  0.1154332  18.198   <2e-16 ***
# fit:Location       -0.0009919  0.0001060  -9.356   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#           Estimate Std. Error z value Pr(>|z|)/2    
# Recipient 0.037721   0.003751   10.06     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1


####Combine Donor and Recipient into one random effect, `Parental`. The estimates are of similar magnitude 
modmat.Donor <- model.matrix(~ 0 + fit:Donor, BB23_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, BB23_aster)
modmat.donorrecipient <- cbind(modmat.Donor,modmat.Recipient)


rout_BB23 <- reaster(resp ~ fit + varb, list(Parental = ~ 0 + modmat.donorrecipient),pred, fam, varb, id, root, data = BB23_aster)
sout_BB23 <-summary(rout_BB23)
sout_BB23

#Call:
# reaster.formula(fixed = resp ~ fit + varb, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = BB23_aster)
# 
# 
# Fixed Effects:
#                    Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.36400    0.07241  -18.84   <2e-16 ***
# fit                 2.74237    0.08010   34.23   <2e-16 ***
# varbF_plant        -4.44010    0.17039  -26.06   <2e-16 ***
# varbsurv_to_flower  1.09505    0.08912   12.29   <2e-16 ***
# varbtotal_fruits    2.10688    0.11545   18.25   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.026337   0.002894     9.1     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

#model comparison for significance of Location. Significant.
rout_BB23_2 <- reaster(resp ~ fit + varb + fit:Location, list(Parental = ~ 0 + modmat.donorrecipient),
    pred, fam, varb, id, root, data = BB23_aster)
sout_BB23_2 <-summary(rout_BB23_2)
sout_BB23_2

save(rout_BB23_2, file = "processed_data/rout_BB23_2.rdata")

# Call:
# reaster.formula(fixed = resp ~ fit + varb + fit:Location, random = list(Parental = ~0 + 
#     modmat.donorrecipient), pred = pred, fam = fam, varvar = varb, 
#     idvar = id, root = root, data = BB23_aster)
# 
# 
# Fixed Effects:
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.3581065  0.0723793 -18.764   <2e-16 ***
# fit                 2.7734798  0.0800776  34.635   <2e-16 ***
# varbF_plant        -4.2783030  0.1701870 -25.139   <2e-16 ***
# varbsurv_to_flower  1.0891535  0.0890964  12.224   <2e-16 ***
# varbtotal_fruits    2.1009877  0.1154341  18.201   <2e-16 ***
# fit:Location       -0.0009984  0.0001064  -9.386   <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Square Roots of Variance Components (P-values are one-tailed):
#          Estimate Std. Error z value Pr(>|z|)/2    
# Parental 0.027985   0.003038   9.211     <2e-16 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

anova(rout_BB23, rout_BB23_2)

# Analysis of Deviance Table
# 
# Model 1: resp ~ fit + varb, ~0 + modmat.donorrecipient
# Model 2: resp ~ fit + varb + fit:Location, ~0 + modmat.donorrecipient
#   Mod Df Fix Mod Df Rand Mod Dev Df Fix Df Rand Deviance    P-value
# 1          5           1  4479.2                                   
# 2          6           1  4594.3      1       0   115.08 7.5698e-27
```

## Mapping function

```{r}
#following Geyer et al. 2022

library(numDeriv)
fishinv<-solve(sout_BB23_2$fisher)
map.factory <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  modmat <- rout$obj$modmat[1, , , drop = FALSE]
  modmat[ , , "fit:Location"] <- 0
  root <- array(5, dim = dim(modmat)[1:2]) 
  x <- aout$x[1, c(1:5)] # added, 5 is the number of nodes 
  # return map function
    print(x) 
  function (b) {
    stopifnot(is.numeric(b))
    stopifnot(is.finite(b))
    stopifnot(length(b) == 1)
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, 
      newcoef = alpha,
      modmat = modmat, 
      root = root, 
      model.type = "conditional",
      is.always.parameter = TRUE)
  
    xi <- matrix(xi, ncol = nnode)
    # always use drop = FALSE unless you are sure you don't want that
    # here if we omit drop = FALSE and there is only one non-subsampling
    # node, the code will break (apply will give an error)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod) #every individual
    # mu is unconditional mean values for model without subsampling
    # in this application all components mu are the same because no
    # covariates except varb, so just return only one
    mu[1] 
  }
}

map <- map.factory(rout_BB23_2, vars %in% c("F_plant", "closed_fruits"))

#MEAN FITNESS
mf<-map(0)
mf
#15.08927

#VA(W)
vaw <- rout_BB23_2$sigma["Parental"]^2 * grad(map, 0)^2
# get rid of name
vaw <- as.numeric(vaw)
vaw <- vaw * 4
vaw
#103.1416

####FFTNS
fftns <- vaw / mf
fftns
#6.835425

# get SE
map.factory.too <- function(rout, is.subsamp) {
  stopifnot(inherits(rout, "reaster"))
  stopifnot(is.logical(is.subsamp))
  aout <- rout$obj
  stopifnot(inherits(aout, "aster"))
  nnode <- ncol(aout$x)
  if (nnode != length(is.subsamp))
    stop("length(is.subsamp) not the number of nodes in the aster graph")
  alpha <- rout$alpha
  ifit <- which(names(alpha) == "fit")
  if (length(ifit) != 1)
    stop("no fixed effect named fit")
  # modmat for one individual. Taking the first row of modmat- there is one row for each individual
  modmat <- rout$obj$modmat[1, , , drop = FALSE] #this only takes the first row 
  modmat[ , , "fit:Location"] <- 0
  # set root = 5
  root <- array(5, dim = dim(modmat)[1:2])
  x <- aout$x[1, c(1:5)] # added, second number is number of nodes 
  # return map function
  function (balpha) {
    stopifnot(is.numeric(balpha))
    stopifnot(is.finite(balpha))
    stopifnot(length(balpha) == 1 + length(alpha))
    b <- balpha[1]
    alpha <- balpha[-1]
    alpha[ifit] <- alpha[ifit] + b
    xi <- predict(
      aout,
      x = x, # added
      newcoef = alpha,
      modmat = modmat,
      root = root,
      model.type = "conditional",
      is.always.parameter = TRUE
    )
    xi <- matrix(xi, ncol = nnode)
    xi <- xi[,!is.subsamp, drop = FALSE]
    mu <- apply(xi, 1, prod)
    # mu is unconditional mean values for model without subsampling
    mu[1]
  }
}

map.too <- map.factory.too(rout_BB23_2, vars %in% c("F_plant", "closed_fruits"))

#centering donor on zero
balpha.hat <- c(0, rout_BB23_2$alpha)

map.too(balpha.hat)
# 15.08927

all.equal(map(0), map.too(balpha.hat))

#first and second derivatives
g<-grad(map.too, balpha.hat)

#calculates the nxn matrix
h <- hessian(map.too, balpha.hat)

#these are partial derivatives
dmu.db <- g[1]
dmu.dalpha <- g[-1]
d2mu.db.dalpha <- h[1, -1]

# give names to the estimators in our formulas
mu.hat <- map.too(balpha.hat)
nu.hat <- rout_BB23_2$nu["Parental"]

#FFTNS SE
#calculations for the gradient vector of the FFTNS prediction with respect to the parameters of the models (fixed effects and variance components) 
#zero removed
dfftns <- c(- 4 * nu.hat * dmu.dalpha * dmu.db^2 / mu.hat^2 +
8 * nu.hat * d2mu.db.dalpha / mu.hat, 4 * dmu.db^2 / mu.hat)

#And apply the delta method.
fftns.se <- t(dfftns) %*% fishinv %*% dfftns
fftns.se <- sqrt(as.vector(fftns.se))
fftns.se
#1.661846

#ADDITIVE GENETIC VARIANCE SE
#calculate the gradient vector VA(W) with respect to the parameters of the model using these formula
dvaw <- c(8 * nu.hat * d2mu.db.dalpha, 4 * dmu.db^2)

#And then apply the delta method to get standard errors
vaw.se <- t(dvaw) %*% fishinv %*% dvaw
vaw.se <- sqrt(as.vector(vaw.se))
vaw.se
#22.56387

#MEAN FITNESS SE
dmf <- c(dmu.dalpha, 0)
mf.se <- t(dmf) %*% fishinv %*% dmf
mf.se <- sqrt(as.vector(mf.se))
mf.se
#1.262928
```

## VA(W) bootstrap SE — Year 2 (eval=FALSE; set TRUE to run, ~30 min for B=50)
```{r vaw-bootstrap-BB23, eval=FALSE}
# #3: Donor-resample bootstrap for VA(W) SE at BB Year 2.
# Year 2 sigma^2(Parental) is two orders of magnitude larger than at the
# other site-years, which pushes the delta-method linearisation into a
# regime where its SE is unreliable. The bootstrap recomputes VA(W) on
# B donor-resampled datasets and gives a SE that does not depend on local
# linearity around sigma = 0.
set.seed(42)
B_boot <- 50
donors_BB23 <- levels(droplevels(as.factor(BB_23$Donor)))

vaw_boot_BB23 <- numeric(B_boot)
for (b_i in seq_len(B_boot)) {
  boot_d <- sample(donors_BB23, replace = TRUE)
  boot_list <- lapply(seq_along(boot_d), function(i) {
    d <- BB23_aster[BB23_aster$Donor == boot_d[i], ]
    d$Donor <- paste0("D", i)
    d
  })
  boot_long <- droplevels(do.call(rbind, boot_list))
  mD <- model.matrix(~ 0 + fit:Donor, boot_long)
  mR <- model.matrix(~ 0 + fit:Recipient, boot_long)
  fit_b <- tryCatch(
    reaster(resp ~ fit + varb + fit:Location,
            list(Parental = ~ 0 + cbind(mD, mR)),
            pred, fam, varb, id, root, data = boot_long),
    error = function(e) NULL)
  if (!is.null(fit_b)) {
    map_b <- map.factory(fit_b, vars %in% c("F_plant", "closed_fruits"))
    nu_b  <- as.numeric(fit_b$nu["Parental"])
    vaw_boot_BB23[b_i] <- nu_b * numDeriv::grad(map_b, 0)^2 * 4
  } else vaw_boot_BB23[b_i] <- NA
  cat("b =", b_i, "VA(W) =", round(vaw_boot_BB23[b_i], 2), "\n")
}
cat("Bootstrap SE:", round(sd(vaw_boot_BB23, na.rm = TRUE), 2),
    "  Delta-method SE:", round(vaw.se, 2), "\n")
cat("Bootstrap 95% CI: [",
    round(quantile(vaw_boot_BB23, 0.025, na.rm = TRUE), 2), ",",
    round(quantile(vaw_boot_BB23, 0.975, na.rm = TRUE), 2), "]\n")
```

## Breeding Values

```{r}
bhat <- rout_BB23_2$b 
bhat.donor<- bhat[grep("Donor", names(bhat))]

#check this is the correct number of donors (49)
length(bhat.donor)

#convert canonical values to mean value parameter values using mapping function
map <- map.factory(rout_BB23_2, vars %in% c("F_plant", "closed_fruits"))
vectorized.map <- Vectorize(map)

#breeding values for donor groups on the mean value parameter scale 
bhat.donor.mu <- vectorized.map(bhat.donor)
bhat.donor.mu<-as.data.frame(bhat.donor.mu)
dmap_BB23 <- Vectorize(function(b) numDeriv::grad(map, b))
bhat.donor.mu$BV_se_approx <- abs(dmap_BB23(bhat.donor)) *
  rout_BB23_2$sigma["Parental"]
write.csv(bhat.donor.mu, "tables/BB23_bhat.Donor.mu.csv")
```


# Bodega Bay Generation 2 2023

```{r}
#subset the dataframe for the progeny generation
BB23_aster_b<-subset(BB_23, Gen=="G2")
BB23_aster_b<-droplevels(BB23_aster_b)
levels(BB23_aster_b$Gen)
```

## Set up aster

```{r}
#establish nodes as vars
vars<-c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

BB23_aster_b <- as.data.frame(BB23_aster_b)

#reshape to long form
BB23_aster_g2 <- reshape(BB23_aster_b,varying = list(vars), direction = "long", timevar = "varb",times = as.factor(vars), v.names = "resp")

#Check that the reshape worked 
nrow(BB23_aster_g2)
nrow(BB23_aster_b)*length(vars)

#designate fitness node, "filled_seeds"
fit<-grepl("filled_seeds", as.character(BB23_aster_g2$varb))
fit<-as.numeric(fit)
BB23_aster_g2$fit<-fit

#check "filled_seeds" is designated as the fitness variable 
with(BB23_aster_g2, sort(unique(as.character(varb)[fit==0])))
with(BB23_aster_g2, sort(unique(as.character(varb)[fit==1])))

#Add a variable "root" to BB_aster_g2 files where value is 5
BB23_aster_g2 <- data.frame(BB23_aster_g2, root=5)

#give each node a number
pred <- c(0,1,2,3,4)

#designate each node with a distribution. 1= Bernoulli, 2= poisson, subsampling node is bernoulli
fam<-c(1,1,2,1,2)

#show graphical model
foo <- c("root", vars)
pvars <- foo[pred +1]
bar <-data.frame(pvars, vars, sapply(fam.default(), as.character)[fam])
colnames(bar) <- c("pred", "succ", "fam")
print(bar, right = FALSE, row.names = FALSE)
```

## Fixed Effect Aster Model

```{r}
BB23_g2 <- aster(resp ~ fit + varb + fit:Location, pred, fam, varb, id, root, data = BB23_aster_g2) 

summary(BB23_g2, info.tol = 1e-9, se.fit=TRUE)

#Call:
# aster.formula(formula = resp ~ fit + varb + fit:Location, pred = pred, 
#     fam = fam, varvar = varb, idvar = id, root = root, data = BB23_aster_g2)
# 
#                      Estimate Std. Error z value Pr(>|z|)    
# (Intercept)        -1.7104906  0.1076837 -15.884  < 2e-16 ***
# fit                 3.2078982  0.1184935  27.072  < 2e-16 ***
# varbF_plant        -4.5828707  0.2630346 -17.423  < 2e-16 ***
# varbsurv_to_flower  1.2734771  0.1340010   9.503  < 2e-16 ***
# varbtotal_fruits    2.5379858  0.1695248  14.971  < 2e-16 ***
# fit:Location       -0.0006232  0.0001102  -5.653 1.58e-08 ***
# ---
# Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
# 
# Original predictor variables dropped (aliased)
#      varbfilled_seeds 

BB_23_hyp <- subset(BB23_aster_g2, id == 1)
# set Location to 0
BB_23_hyp$Location <- 0
# set resp to 1
BB_23_hyp$resp <- 1


# predict conditional fitness estimates for this hypothetical individual
BB_23_hyp.p <- predict(BB23_g2, varvar = varb, idvar = id, root = root,
  newdata = BB_23_hyp, model.type = "conditional")

# take the product of these conditional estimates, EXCEPT for the sub sampling nodes (2nd and 4th node)
mu.fit<-prod(BB_23_hyp.p[c(-2,-4)])
mu.fit
#80.67425
```

## Estimate Mean Fitness and SE
```{r}
#use poisson nodes 
#conditional
nnode <- length(vars)
nind <- length(unique(BB_23_hyp$id))
nnode * nind == nrow(BB_23_hyp)

pout.cond <- predict(BB23_g2, newdata = BB_23_hyp, varvar = varb, idvar = id, root = root, model.type = "conditional", is.always.parameter = TRUE, gradient= TRUE)
xi<-pout.cond$fit

class(xi)
length(xi)==nind*nnode

xi <- matrix(xi, nrow = nind)
colnames(xi) <- vars
xi

#      surv_to_flower   F_plant total_fruits closed_fruits filled_seeds
# [1,]      0.4589572 0.6144633      7.87014     0.8527421     4.466942

#unconditional
pout.unco<-predict(BB23_g2, newdata = BB_23_hyp, varvar = varb, idvar = id, root = root, gradient=TRUE)
mu<-pout.unco$fit
mu <- matrix(mu, nrow = nind)
colnames(mu) <- vars
mu

#      surv_to_flower  F_plant total_fruits closed_fruits filled_seeds
# [1,]       2.294786 1.410062     11.09738      9.463204     42.27159

is.seeds<-grep("filled_seeds", vars)
is.fruits<-grep("total_fruits", vars)

is.seeds
is.fruits

#unconditional estimate of total fruits 
mu.fruits<-mu[, is.fruits]

#conditional estimate of total seeds 
xi.seeds <- xi[ , is.seeds]

mu.seeds <- mu.fruits * xi.seeds
mu.seeds
# Total_fruits
#    49.57136

#make a  function
foo <- function(x) {
# x is xi and mu strung out as one vector
xi <- x[1:length(xi)]
mu <- x[- (1:length(xi))]
xi <- matrix(xi, nrow = nind)
mu <- matrix(mu, nrow = nind)
mu.fruits <- mu[ , is.fruits]
xi.seeds <- xi[ , is.seeds]
mu.seeds <- mu.fruits * xi.seeds
}

#vector of conditional and unconditional estimates for each node 
ximu <- c(xi, mu)
all.equal(foo(ximu), mu.seeds)

foo(ximu)
mu.seeds
#total_fruits
#   49.57136

library(numDeriv)
# R package numDeriv figures out the jacobian matrix for this transformation.
#We also need the jacobian matrix for the transformation from the “coefficients” vector to the vector ximu

jac.foo <- jacobian(foo, ximu)
jac.ximu <- rbind(pout.cond$gradient, pout.unco$gradient)
jac.total <- jac.foo %*% jac.ximu
#Now the delta method says the variance-covariance matrix of all the fitnesses (the vector estimate mu.fit) is JI−1JT, where J is the overall jacobian matrix jar.total and I is Fisher information for the “coefficients” vector
#there are six coefficients from the model
V <- jac.total %*% solve(BB23_g2$fisher) %*% t(jac.total)
#and the standard errors are square roots of the variances (the diagonal elements of V)
se <- sqrt(diag(V))
bar.pois <- cbind(as.numeric(mu.seeds), se)
colnames(bar.pois) <- c("Estimate", "SE")
bar.pois
# Estimate       SE
#49.57136 6.692107
```

```{r}
#BACK OF NAPKIN FITNESS CALCULATION BB 2023 G1

W <- BB23_aster_a %>%
  summarise(
    total_seeds = sum(filled_seeds, na.rm = TRUE),
    W = total_seeds / n()
  )

W
#18.27689


#BACK OF NAPKIN FITNESS CALCULATION BB 2023 G2

W <- BB23_aster_b %>%
  summarise(
    total_seeds = sum(filled_seeds, na.rm = TRUE),
    W = total_seeds / n()
  )

W
#19.97487

```


#GxE Effects

##merge the datasets for GxE analysis
```{r}
library(dplyr)

BB_22_gxe <- BB_22 %>%
  mutate(Year = "2022") %>%
  dplyr::select(Location, Donor, Recipient, sample_ID, surv_to_flower, F_plant,
         total_fruits, closed_fruits, filled_seeds, Year)

BB_23_gxe <- BB_23 %>%
  mutate(Year = "2023") %>%
  dplyr::select(Location, Donor, Recipient, sample_ID_SEG, surv_to_flower, F_plant,
         total_fruits, closed_fruits, filled_seeds, Year) %>%
  rename(sample_ID = sample_ID_SEG)

BB_G1_both <- bind_rows(BB_22_gxe, BB_23_gxe) %>%
  mutate(
    Year = factor(Year),
    Donor = factor(Donor),
    Recipient = factor(Recipient)
  )
```

##set up aster
```{r}
vars <- c("surv_to_flower", "F_plant", "total_fruits", "closed_fruits", "filled_seeds")

BB_G1_both <- as.data.frame(BB_G1_both)

BB_G1_aster <- reshape(
  BB_G1_both, varying = list(vars), direction = "long", timevar = "varb", times = as.factor(vars), v.names = "resp")

#####Check that the reshape worked 
#If it did the number of rows will be the same 
nrow(BB_G1_aster) #4790
nrow(BB_G1_both)*length(vars) #4790

fit <- grepl("filled_seeds", as.character(BB_G1_aster$varb))
fit <- as.numeric(fit)
BB_G1_aster$fit <- fit

#set up root that is year specific

BB_G1_aster$root <- ifelse(BB_G1_aster$Year == "2022", 8, 5) #This is important bc the root changes in 2023 vs 2022!

pred <- c(0,1,2,3,4)
fam  <- c(1,1,2,1,2)

#check that the distribution you assigned is correct
sapply(fam.default(), as.character)[fam]

#show graphical model
foo <- c("root", vars)
pvars <- foo[pred +1]
bar <-cbind(pvars, vars)
colnames(bar) <- c("pred", "succ")
```

##set up parental and random effects
```{r}
#parental effect
modmat.Donor <- model.matrix(~ 0 + fit:Donor, BB_G1_aster)
modmat.Recipient <- model.matrix(~ 0 + fit:Recipient, BB_G1_aster)
modmat.donorrecipient <- cbind(modmat.Donor, modmat.Recipient)

#GxE random effect
modmat.parental.year <- model.matrix(~ 0 + fit:Donor:Year, BB_G1_aster)
modmat.recipient.year <- model.matrix(~ 0 + fit:Recipient:Year, BB_G1_aster)
modmat.parentalyear <- cbind(modmat.parental.year, modmat.recipient.year)
```

##run both models
```{r}
#fit model WITHOUT GxE
rout_no_gxe <- reaster(
  resp ~ fit + varb + fit:(Year + Location),
  random = list(Parental = ~ 0 + modmat.donorrecipient),
  pred = pred,
  fam = fam,
  varvar = varb,
  idvar = id,
  root = root,
  data = BB_G1_aster
)

#fit model WITH GxE

rout_gxe <- reaster(
  resp ~ fit + varb + fit:(Year + Location),
  random = list(
    Parental = ~ 0 + modmat.donorrecipient,
    ParentalYear = ~ 0 + modmat.parentalyear
  ),
  pred = pred,
  fam = fam,
  varvar = varb,
  idvar = id,
  root = root,
  data = BB_G1_aster
)
```


##model comparison
```{r}
##################################################################
#Likelihood ratio test for effect of Donor X Year interaction
##################################################################
gxy_lrt <- anova(rout_no_gxe, rout_gxe)
gxy_lrt
summary(rout_gxe)

# Boundary-corrected p-value (Self & Liang 1987): H0: sigma^2(ParentalYear) = 0
# lies on the boundary, so the reference distribution is 50:50 chi^2_0 + chi^2_1.
# Halving the standard LRT p-value gives the correct one-sided boundary test.
cat("\nBoundary-corrected G×Y p:", gxy_lrt$`Pr(>Chisq)`[2] / 2, "\n")
```


## Between-year sire correlation with shrinkage diagnostic (#6)
```{r sire-correlation-shrinkage}
# Load both years of BLUPs for this site. The Pearson r is influenced by
# year-specific shrinkage asymmetry (sigma differs between years), which
# can create apparent rank changes even when true breeding values are
# stable. Spearman r is more robust. We also report sigma_Y1/sigma_Y2 as
# a shrinkage-asymmetry diagnostic.
site_code <- sub("Aster6_([A-Za-z]+)\\.Rmd", "\\1",
                 knitr::current_input())

# resolve to 2-letter code
site_map <- c(Angelo="AC", BlueOak="BO", BodegaBay="BB", Hastings="HR")
site <- site_map[site_code]

bv22_path <- paste0("tables/", site, "22_bhat.Donor.mu.csv")
bv23_path <- paste0("tables/", site, "23_bhat.donor.mu.csv")

bv22 <- read.csv(bv22_path)
bv23 <- read.csv(bv23_path)
names(bv22)[1:2] <- c("raw", "BV_Y1")
names(bv23)[1:2] <- c("raw", "BV_Y2")
bv22$Donor <- sub(".*Donor", "", bv22$raw)
bv23$Donor <- sub(".*Donor", "", bv23$raw)
bv_both <- merge(bv22[, c("Donor","BV_Y1")],
                 bv23[, c("Donor","BV_Y2")], by = "Donor")

pearson  <- cor.test(bv_both$BV_Y1, bv_both$BV_Y2, method = "pearson")
spearman <- cor.test(bv_both$BV_Y1, bv_both$BV_Y2, method = "spearman",
                     exact = FALSE)

# Sigma values were computed earlier in this script
sig_y1 <- rout_BB22_2$sigma["Parental"]
sig_y2 <- rout_BB23_2$sigma["Parental"]

cat("
--- Between-year sire correlation:", site, "---
")
cat("n donors with both years:", nrow(bv_both), "
")
cat("Pearson  r =", round(pearson$estimate, 3),
    " p =", round(pearson$p.value, 4), "
")
cat("Spearman rho =", round(spearman$estimate, 3),
    " p =", round(spearman$p.value, 4), "
")
cat("sigma_Y1:", round(sig_y1, 5),
    "  sigma_Y2:", round(sig_y2, 5),
    "  ratio Y1/Y2:", round(sig_y1/sig_y2, 2), "
")
cat("(ratio >> 1 or << 1 indicates shrinkage asymmetry that biases Pearson r)
")
```

